\part*{Childhood}
\makeoddhead{modruled}{}{\scshape Childhood}{}
\markboth{Childhood}{}

\chapter{Karl Ivanych, the Tutor} %c1

On the 12th of August, 18\ldots{}, exactly the third day after my birthday, when I turned ten and received some marvelous gifts, Karl Ivanych woke me up at seven o'clock in the morning, hitting the wall just above my head with a flyswatter made from sugar paper and a stick. He did this so carelessly that he hit the icon of my angel, which was hanging from the oak headboard, and the dead fly fell right on my head.\footnote{By angel, Tolstoy means the narrator's saintly namesake. --- Trans.} I stuck my nose out from under the cover, stopped the icon, which was still swinging, with my hand, knocked the dead fly onto the floor, and threw an angry look at Karl Ivanych, my eyes still clouded with sleep. He was wearing a motley cotton dressing-gown, belted with a belt from the same material, a red knit cap with a tassel, and soft goat-skin \todo{?} boots, and he continued to walk along the walls, looking for targets and slapping with his flyswatter.

``Fine,'' I thought, ``I \emph{am} small, but why does he keep bothering me? Why doesn't he swat flies by Volodya's bed? There are so many over there! No, Volodya is older than me; I'm the youngest: that's why it's me he torments. That's what he thinks about all day,'' I whispered, ``how to make trouble for me. He can see very well that he woke me up and frightened me, but he acts like he doesn't notice\ldots{} what a disgusting person! His robe, his little hat, and that tassel---disgusting!'' %narr

At the same time I was mentally expressing my annoyance with Karl Ivanych, he went over to his bed, looked at his watch, which was hanging over it in a beaded slipper, hung the flyswatter up on its nail, and, obviously in a very pleasant mood, turned to us.

\textit{``Auf, Kinder, auf!\ldots{} s'ist Zeit. Die Mutter ist schon im Saal,''} he shouted in his kind German voice, then he went over to me, sat by my feet, and took his snuff box out of his pocket.\footnote{Up, children, up! 'Tis time. Mother is already in the hall. \textit{(Ger.)}} I pretended I was still sleeping. Karl Ivanych took snuff, wiped his nose, snapped his fingers, and then started in on me. Laughing, he started to tickle the soles of my feet. \textit{``Nu, nun, Faulenzer,''} he said.\footnote{Well, now, lazy boy! \textit{(Ger.)}} %karl

However much I dreaded the tickling, I did not jump out of bed and did not answer him, instead hiding my head deeper under the pillows, kicking my feet as hard as I could and doing everything I could to keep from laughing.

``He's so kind and loves us so much, and I was thinking such awful things about him!'' %narr

I was annoyed both at myself and at Karl Ivanych, wanting to laugh and wanting to cry: my nerves were unsettled.

\textit{``Ach, lassen sie, Karl Ivanych!''}~I shouted with tears in my eyes, sticking my head out from under the pillows.\footnote{Leave me alone, Karl Ivanych! \textit{(Ger.)}} %narr

Karl Ivanych seemed surprised, left the soles of my feet in peace, and started asking with concern: what was it? had I had a nightmare?\ldots{} his kind German face and the sympathy with which he tried to guess the reason for my tears made them flow all the more abundantly: I was ashamed, and I could not understand how just a minute before I could not love Karl Ivanych and find his robe, hat, and tassel so disgusting; now, on the contrary, all of this seemed extremely endearing, and even the tassel seemed like a clear indication of his kindness. I told him that I was crying because I had had a nightmare---as if \textit{maman} had died and they were taking her to the grave.\footnote{Tolstoy uses French \textit{papa} and \textit{maman} throughout for the narrator's parents. Because they are used so extensively, I have left them untranslated but also unitalicized except for the first occurrence. --- Trans.} I made all this up, because I absolutely could not remember what I had dreamed about that night; but when Karl Ivanych, touched by my story, started to comfort and soothe me, I started thinking I had indeed had that nightmare, and my tears started flowing for another reason. 

When Karl Ivanych left me, and I, sitting up a little in bed, started to pull my stockings \todo{?} onto my little legs, my tears had subsided a little, but the dark thoughts brought on by my imagined dream did not leave me. Uncle \todo{need a footnote...} Nikolai came in---a small, neat little man, always serious, exact, respectable, \todo{?} and a great friend of Karl Ivanych. He brought our dresses and shoes: boots for Volodya, for me still an intolerable pair of children's shoes with bows. \todo{specific term for shoes?} I was ashamed to cry in front of him; the morning sun was shining through the window besides, and Volodya, mimicking Marya Ivanovna (our sisters' governess), was laughing so happily and sonorously, standing over the wash stand, that even serious old Nikolai, a towel over his shoulder, with soap in one hand and a basin in the other, smiling, said:

``That's enough, Vladimir Petrovich, kindly wash up.'' %unclenikolai

I cheered up completely.

\textit{``Sind sie bald fertig?''} Karl Ivanych's voice came from the classroom.\footnote{Are you almost ready? \textit{(Ger.)}} %karl

His voice was strict and no longer had the same tone of kindness that had touched me and brought me to tears. In the classroom, Karl Ivanych was an entirely different person: he was a schoolmaster. I dressed quickly, washed up, and, with a brush in my hand, smoothing my wet hair, I answered his call.

Karl Ivanych, with glasses on his nose and a book in his hand, sat in his usual place, between the door and the window. To the left of the door were two little shelves: one was ours, the children's, the other was Karl Ivanych's, his \emph{personal} shelf. On ours were all sorts of books---educational and non-educational: some were standing up, others were lying down. Only the two large volumes of \textit{Histoire des voyages} in red covers rested sedately against the wall;\footnote{History of the Voyages. \textit{(Fr.)}} \todo{Voltaire?} then there were long, thick, large, and small books---covers without books and books without covers; we used to push and shove them in there before recess, when we were told to put the library, as Karl Ivanych loudly called that shelf, in order. The collection of books on his \emph{personal} shelf, if not so large as the one on ours, was even more varied in composition. From those books, I remember three: a German brochure on the manuring of cabbage gardens---without a cover, one volume containing a history of the Seven Years' War---in parchment, burnt at one corner, and a full course on hydrostatics. Karl Ivanych spent the greater part of his time reading, even ruining his eyes in the process; but other than those books and the \textit{Northern Bee}, he read nothing else.

Among the items lying on Karl Ivanych's shelf, I am constantly reminded of one in particular. It was a little circle of cardboard, set on a wooden base to which the cardboard circle was attached by way of a few pins. On this circle, a picture was pasted, a caricature of a lady and a hairdresser. Karl Ivanych was very good at pasting things together and had devised the circle himself and made it to protect his weak eyes from bright light.\todo{Run on ok?}

I can see him before me now a long figure in a cotton dressing-gown with sparse grey hair sticking out from his little red hat. He sits by a little table, on which the circle with the hairdresser is standing, casting a shadow over his face; in one hand he holds a book, the other rests on the arm of the chair; near him lies his watch, with a painted hunter on the dial, a checkered handkerchief, a round black snuff box; a green eyeglass case, a set of tongs on a tray. \todo{I think this is right, not translated by the Maudes!} Every item was so sedately, so exactly placed that by their good order alone one could conclude that Karl Ivanych's conscience was clean and his soul at peace.

At times, after we had had enough of running around downstairs in the hall, I would sneak up to the classroom on tiptoe---Karl Ivanych would be sitting there alone in his chair, reading one of his favorite books with a peaceful and sublime expression. Sometimes I would find him in a moment when he was not reading: his glasses would be at the very end of his aquiline nose, his blue eyes, half-closed, would be staring off into the distance with a peculiar gaze, and his lips would be smiling sadly. It would be quiet; the only sounds would be his even breathing and the ticking of his watch with the hunter.

At times he would not notice me, and I would stand at the door and think: poor, poor old man! There are so many of us, we are playing, we are enjoying ourselves, and he is all by his lonesome, \todo{neologism, ok?} and he has no one to show him affection. It is true what he says, that he is an orphan. And the story of his life is so terrible! I remember when he told it to Nikolai---it would be terrible to be in his position! And I feel such pity for him that I would at times come up to him, grab his hand, and say, \textit{``Lieber Karl Ivanych!''} He loved it when I talked to him like that; he would always look at me with affection, and I could tell that he was deeply moved.

On the second wall hung maps, nearly all of them tattered, but skillfully mended by Karl Ivanych's hand. On the third wall, in the middle of which was the door leading downstairs, on one side hung two rulers: one, ragged, was ours, the other, a new one, was his \emph{personal} ruler, used more for discipline than for drawing lines; on the other side was a blackboard on which our serious offences were marked with circles and our minor ones with little crosses. To the left of the board was a corner, where they made us kneel.

How I remember that corner! I remember the flap of the stove, the vent in that flap, and the noise that it made when you moved it. At times I would be kneeling there in the corner, kneeling, kneeling, so that my knees and back hurt, and I would think: Karl Ivanych has forgotten about me: I suppose he is sitting there peacefully in his soft chair and reading his hydrostatics---and what about me? And I would start to open and shut the vent or pick plaster off from the wall; but if too large a piece suddenly fell on the ground with a loud noise, well, the fear alone was worse than any punishment. I would glance over at Karl Ivanych---and he would be sitting quietly with a book in his hand, as if he had not noticed a thing.

In the middle of the room stood a table covered with ragged oilcloth, the corners visible underneath ragged with the cuts of our pen knives. Around the table were a few stools, unfinished but worn smooth from long use. The final wall was taken up by three little windows. This was the view from them: right under the windows was a road, every potholes, every pebble, every rut of which was long since known and dear to me; beyond the road was a little park full of trimmed lilac, behind which a wicker fence could be seen; through the trees, a meadow was visible, with a barn to one side of that and the forest opposite; far away in the forest, the warden's little hut was visible. From the window, to the right, part of the terrace was visible, where the grown-ups would usually sit until dinner. At times, while Karl Ivanych was correcting our dictation, I would look out that side, see mother's black hair, someone's back, and hear vague conversations and laughter from there; then I would get annoyed that I could not be there and think: ``When will I be a grown-up, stop studying, and be able to sit with the people I love instead of doing dialogues?'' My annoyance would turn into sadness and, God knows why, or what I was thinking about, but I would not hear a word of Karl Ivanych, who was angry about my mistakes. %narr

Karl Ivanych took off his robe, put on his blue tail coat with pleats and folds \todo{?} on the shoulders, straightened his tie in front of the mirror, and guided us downstairs to say good morning to mother.

\chapter{Maman} %c2

Mother sat in the drawing room and was pouring tea; with one hand she held the teapot, and with the other the tap of the samovar, from which water was flowing through the top of the teapot and onto the tray. But although she was staring intently, she did not notice this, nor did she notice that we had come in.

So many memories of the past surface when you try to resurrect features of a loved one that you can see them only vaguely through those memories, clouding your vision like tears. Those are tears of the imagination. When I try to remember mother as she was at that time, I see only her brown eyes, which always expressed kindness and love equally, the mole on her neck, a little below the place where her little curly hairs began, her embroidered white collar, her dry, tender hand, which caressed me so often and which I often kissed; but the overall picture slips eludes me.

To the left of the sofa stood an old English piano; in front of the piano sat my dark-looking sister, Lyubochka, her pink fingers, just washed with cold water, playing Clementi's etudes with visible effort. She was eleven years old; she went around in a short linen dress and white, lace-trimmed pants and could manage octaves only in arpeggio. Near her, half-turned away, sat Marya Ivanovna in a bonnet with pink ribbons, in a light blue caraco jacket and with a red, angry face, which took on an even stricter expression as soon as Karl Ivanych walked in. She looked at him menacingly and, not answering his bow, continued, tapping her foot, to count: \textit{un, deux, trois, un, deux, trois} louder and more demandingly than before.

Karl Ivanych, not paying even the slightest attention to this, as usual, went directly to mother's hand with a German greeting. She came to her senses, shook her head, as if by this gesture trying to drive away unpleasant thoughts, gave her hand to Karl Ivanych and kissed him on his wrinkled temple, at the same time he kissed her hand:

\textit{``Ich danke, lieber Karl Ivanych,''} and, continuing to speak in German, she asked:\footnote{My thanks, dear Karl Ivanych. \textit{(Ger.)}} %maman

``Did the children sleep well?'' %maman

Karl Ivanych was deaf in one ear, and now, because of the noise from the piano, he could hear nothing at all. He stooped closer to the sofa, leaned with one arm on the table, supporting himself with one leg, and, with a smile that seemed to me the height of sophistication, lifted his hat a little off his head and said:

``You will forgive me, Natalya Nikolayevna?'' %karl

Karl Ivanych, so that his bare head did not catch cold, never took off his little red hat, but he asked forgiveness for this each time he came into the drawing room.

``Put it back on, Karl Ivanych\ldots{} I am asking you, did the children sleep well?'' asked maman fairly loudly, moving closer to him. %maman

But he again heard nothing, covered his bald spot with the little red hat, and smiled even more sweetly.

``Stop for a minute, Mimi,'' said maman to Marya Ivanovna with a smile: ``we can't hear anything.'' %maman

When mother smiled, though her face was already pretty, she became incomparably more beautiful, and everything around her seemed to become brighter. If in the heaviest moments of my life I could have just caught a glimpse of that smile, then I would never have known what sorrow is. It seems to me that the smile alone holds what we call beauty: if a smile adds charm to a face, then the face is very fine; if it does not change, then it is ordinary; and if it ruins the face, then it is ugly.

Having greeted me, maman took my head in both hands and leaned it back, then looked at me fixedly and said:

``Did you cry today?'' %maman

I did not answer. She kissed my eyes and asked in German:

``What were you crying about?'' %maman

When she talked with me in this friendly way, she always spoke in that language, which she had mastered to perfection.

``It was a dream, maman, that's why I was crying,'' I said, remembering in all its detail my made-up dream and shuddering involuntarily at the thought. %narr

Karl Ivanych confirmed my words but was silent about the dream. Changing the subject to the weather---a conversation which Mimi also took part in---maman put six pieces of sugar for some of the more valued servants, stood up and went over to the lace frame that stood by the window.

``Well, go to \textit{papa} now, children, and tell him he needs to come see me before he goes out to the barn.'' %maman

The music, the counting, and the menacing glances began again, and we went to papa. Walking through another room, which from my grandfather's days was still called the footman's pantry, we went into his study.

\chapter{Papa} %c3

He was standing by his desk and, pointing at some envelopes, papers, and piles of money, speaking heatedly, and explaining something angrily to the steward, Yakov Mikhaylov, who was standing in his usual place between the door and the barometer, his hands placed behind his back, and was quickly moving his fingers back and forth.

The angrier papa got, the faster the fingers moved: likewise, when papa stopped, the fingers stopped as well; but when Yakov himself started talking, the fingers became extremely agitated and flew around desperately in every direction. It seemed to me that one could guess Yakov's secret thoughts by their movements; his face was always calm---it expressed an awareness of the merits of his case, but also of his subordinate position, in other words: I am right, however, as you like it!

Seeing us, papa merely said:

``Wait, just a moment.'' %papa

And he nodded his head at the door, so that one of us would shut it.

``Oh, merciful heavens! What is with you today, Yakov?'' he continued saying to the steward, twitching his shoulder (he had a habit of doing that). ``This envelope, with 800 rubles enclosed\ldots{}'' %papa

Yakov moved beads on the counting frame, sliding over 800, and turned his gaze on some indeterminate point, waiting for what would come next.

``\ldots{} for housekeeping expenses in my absence. Understand? You should get 1,000 rubles from the mill\ldots{} Yes or no? You should get 8,000 back from the Treasury bonds; by your estimates we can sell 7,000 pounds of hay---I will put it out for 45 kopecks---so you should get 3,000; so how much money do you get in the end? 12,000\ldots{} Yes or no?'' %papa

``Yes, exactly, sir,'' said Yakov. %yakov

But from the quickness with which his fingers moved, I could tell he wanted to object; papa stopped him:

``Well, from that money you will send 10,000 to the Council for Petrovskoye. Then the money that is at the bank,'' continued papa, while Yakov mixed up the count and put up 21,000 instead of 12,000, ``you will bring to me and show as an expense as of this date.'' Yakov mixed up the count and turned over the counting-frame, apparently showing by this movement that the 21,000 would be spent. ``This envelope with the money you will send in my name to the address listed.'' %yakov

I was standing close to the desk and looked at the inscription. On the envelope was written: ``To Karl Ivanovich Mauer.''

Apparently noticing that I had read something that I did not need to see, papa put his hand on my shoulder and with an easy movement showed me away from the desk. I did not know whether it was a caress or a reproach; in any case, I kissed the large, sinewy hand that was lying on my shoulder.

``Yes, sir,'' said Yakov. ``And what orders are there for the Khaborovka money?'' %yakov

Khabarovka was mother's village.

``Leave it in the bank and do not use them for anything, under any circumstances, without my orders.'' %papa

Yakov was silent for a few seconds; then suddenly his fingers began to whirl around with double the speed, and he, changing his expression from the one of dull obedience with which he listened to his master's orders to his more characteristic expression of sharp cunning, moved the counting frame toward him and began to speak:

``Permit me to report, Pyotr Aleksandrych, that it will not be possible to pay even the Council on time as you wish. You were so kind as to say,'' he continued after a pause, ``that we should receive money from the bonds, from the mill, and from the hay\ldots{}'' Counting out all of these items, he slid beads over to show each one. ``But I am afraid we may be mistaken in our estimates,'' he added, falling silent for a while and looking gravely at papa. %yakov

``Why?'' %papa

``If you would be so kind as to see: regarding the mill, the miller has come to me twice already to ask for a deferment and swore by Christ our Lord that he did not have the money\ldots{} And he is actually here now: perhaps you would like to speak with him yourself?'' %yakov

``What does he have to say?'' asked papa, shaking his head to show that he did not want to speak with the miller. %papa

``Isn't it clear? He says there was nothing to mill, and what little money there was, he had to put it all into the dam. So, if we take him off the mill, \emph{sir}, do you think we will make our estimates then? You were so kind as to speak about the bonds, well, I believe I already reported to you that once we put our money in them, we could not get it back soon. The other day, I sent a cart of flour to Ivan Afanasyich in the city, along with a note about this matter: he answered the same as before, I am happy to make an effort for Pyotr Aleksandrych, but the matter is not in my hands, and that everything suggests it will be two months or more until our check comes. You were so kind as to speak about the hay, let us assume that it can sell for 3,000\ldots{}'' %yakov

He slid 3,000 over on the counting frame and was silent for a minute, looking first at the counting frame, then into papa's eyes, with an expression that said:

``You can see for yourself how little this is! And we'll take a loss on the hay in any case if we sell it right now, you know that yourself\ldots{}'' %yakov

It was clear that had a large reserve of such arguments left; it must have been for this reason that papa interrupted him.

``I am not going to change my orders,'' he said, ``but if there is, in fact, any delay in receiving this money, then there's nothing that can be done, take whatever is needed from the Khaborovka money.'' %papa

``Yes, sir.'' %yakov

By Yakov's fingers and the expression on his face, it was obvious that this last order gave him great pleasure.

Yakov was a serf, an extremely diligent and devoted person; like all good stewards, he was tightfisted in the extreme on his master's behalf and had the strangest conceptions about his master's interests. He was perpetually looking after profit for his master's estate at the expense of his mistress's estate, always trying to prove that it was unavoidable to spend the revenues from her property on Petrovskoye (the village in which we lived). At the present moment, he was triumphant, because he had succeeded perfectly in this.

Having greeted us, papa told us what lazybones we had become out in the country, that we were no longer little, and that it was time for us to study seriously.

``You already know, I think, that I am going to Moscow tonight, and I am taking you with me,'' he said. ``You are going to live at your grandmother's, and maman and the girls will stay here. And you know that only one thing will make me pleased---and that is to hear that you are studying hard and that they are satisfied with you there.'' %papa

Although the preparations that had been obvious for a few days had prepared us for something unusual to happen, this news nevertheless affected us terribly. Volodya turned red and with a quivering voice gave him mother's message.

``Then this is what my dream was predicting!'' I thought. ``Please God, don't let anything worse happen.'' %narr

I was very, very sad for mother, but, at the same time, the thought that we had become grownups made me happy.

``If we are going today, then surely there won't be any classes: that is nice!'' I thought. ``But I am sad for Karl Ivanych. They will probably let him go, or else they wouldn't have made that envelope for him\ldots{} It would be better to have to study for a century but not go away, not leave mother behind, and not treat poor Karl Ivanych so badly. He is already so unhappy!'' %narr

These thoughts flashed through my head; I did not move from the spot and looked fixedly at the black bows on my shoes.

Having said a few words with Karl Ivanych about the falling barometer and having ordered Yakov not to feed the dogs, so he could ride out after dinner on a fairwell hunt with the young hounds, papa, against my expectations, sent us to study, consoling us, however, with a promise to take us on the hunt.

On the way upstairs I ran out onto the terrace. Near the doors, lying in the son and squinting her eyes, lay father's favorite borzoi dog, Milka.

``Milochka,'' I said, caressing her and kissing her snout, ``we are going today; goodbye! We'll never see each other again.'' %narr

I was deeply affected by this and cried.

\chapter{Classes} %c4

Karl Ivanych was already in a bad mood. This was obvious from his furrowed brow, and from the way he had thrown his frock coat in the bureau, \todo{?} and how he angrily belted up his robe, and how forcefully he had dug his nail into the place in the dialogue book where the passage that we were supposed to memorize ended. Volodya made a decent effort at studying; I, on the other hand, was so upset that I could not do anything at all. For a long time, I sat gazing senselessly at the dialogue book, but because of the tears welling up in my eyes at the thought of our imminent parting, I could not read; when the time came for me to repeat them to Karl Ivanych, who, squinting, listened to me (this was a bad sign), I came to the place where one person says, \textit{``Wo kommen sie her?''} and the other answers: \textit{``ich komme vom Kaffe-Hause,''} and I could not hold back my tears any longer, unable because of my sobbing to pronounce the words: \textit{``Haben sie die Zeitung nicht gelesen?''}\footnote{Where are you coming from? I am coming from the cafe. Have you read the newspaper? \textit{(Ger.)}} When we turned to penmanship, I left so many blots on the page because of the tears that I let fall that it was like I was writing with water on parcel paper.

Karl Ivanych got angry, made me kneel in the corner, said over and over that I was being obstinate, that I was playing a farce (this was his favorite phrase), threatened me with the ruler, and demanded that I ask his pardon when I could barely get out a word for crying; finally, seemingly feeling the injustice of what he was doing, he went into Nikolai's room and slammed the door.

From the classroom, I could hear the conversation in Uncle Nikolai's room.

``You heard, Nikolai, that the children go to Moscow?'' said Karl Ivanych, entering the room. %karl

``I heard, of course.'' %unclenikolai

Nikolai must have been about to stand up, because Karl Ivanych said, ``Sit, Nikolai!'' and right after that shut the door. I came out of the corner and went over to the door to eavesdrop.

``No matter what kind things you do to people, no matter how attached you are, you obviously cannot be expecting gratitude, eh, Nikolai?'' said Karl Ivanych with feeling. %karl

Nikolai, sitting by the window, repairing shoes, nodded his head.

``I am living in this house twelve years and can say before God, Nikolai,'' continued Karl Ivanych, lifting his eyes and his snuff box toward the ceiling, ``that I have loved them and helped them more than as if they were my own children. You remember, Nikolai, when Volodyenka had the fever, you remember how I sat for nine days by his bed, not closing my eyes. Yes: then I was nice, kind Karl Ivanych, then I was needed; but now,'' he added, smiling ironically. ``Now \emph{the children have become grown ups: they need to study seriously.} As if they don't study anything here, right, Nikolai?'' %karl

``Of course they do,'' said Nikolai, putting down his stitching awl and spreading out the thread with both hands. %unclenikolai

``Well, now I have become not needed, they need to throw me out; and where are the promises? where is the gratitude? I love and respect Natalya Nikolayevna, Nikolai,'' he said, placing his hand on his chest, ``but what does she have to do with it\ldots{}? They don't care about her desires in this house, that's it.'' At this he threw a scrap of leather onto the floor with a significant gesture. ``I know whose tricks these are and why I have become not needed: because I am not always flattering and conniving like \emph{certain people}. I am used to always and in front of everyone telling the truth,'' he said proudly. ``Forget about them, for God's sake! They are not getting any richer because I will not be here, and I will find my bite of bread, if God is merciful\ldots{} Don't you think, Nikolai?'' %karl

Nikolai lifted his head and looked at Karl Ivanych as if to ascertain whether he really could find his bite of bread---but he said nothing.

Karl Ivanych went on in this vein for a long time: he talked about how they knew how to value his services much better in some general's house where he lived before (it was very painful for me to hear that), talked about Saxony, about his parents, about his friend, the tailor Sch\"onheit, etc., etc.

I empathized with him in his sorrow, and it pained me that father and Karl Ivanych, whom I loved almost equally, did not understand one another; I again went into the corner, got down on my knees, and debated with myself about how to re-establish harmony between them.

Returning to the classroom, Karl Ivanych told me to stand up and prepare my notebook for writing from dictation. When everything was ready, he lowered himself majestically into his chair and, in a voice that seemed to come from a great depth, began to dictate the following: \textit{``Von al-len Lei-den-schaf-ten die grau-samste ist\ldots{} haben sie geschrieben?''}\footnote{Of all the passions, the cruelest is\ldots{} Have you written that? \textit{(Ger.)}} Here he stopped, slowly took snuff, and continued with new strength: \textit{``die grausamste ist die Un-dank-bar-keit\ldots{}''}\footnote{The cruelest is Ingratitude\ldots{} \textit{(Ger.)}} Waiting for the continuation, having written the last word, I looked at him. %karl

\textit{``Punctum,''} he said with a barely noticeable smile and gestured for us to give him our notebooks.\footnote{Period. \textit{(Lat.)}} %karl

He read this aphorism several times with different intonations and with an expression of the greatest pleasure, expressing as it did his mostly deeply felt thought; then he gave us a history lesson and sat down by the window. His face was not sullen as before; it expressed the contentment of a person who had suitably avenged himself for a wrong imposed on him.

It was a quarter till the hour; but Karl Ivanych, it seemed, was not thinking about letting us go: he was constantly giving us new lessons. Boredom and our appetites increased in equal measure. With the greatest impatience I followed the signs that suggested dinner was near. First a servant girl with a washcloth coming to wash the plates, then the sound of dishes in the dining room, the table being moved and the chairs begin dragged into place, then Mimi with Lyubochka and Katyenka (Katyenka was Mimi's twelve year old daughter) coming in from the garden; but Foka was nowhere to be seen---Foka, the butler, who always came in and announced that the food was ready. Only then could we throw down our books and, paying no attention to Karl Ivanych, run downstairs.

Just then we heard steps on the stairs; but it was not Foka! I had learned his gait and could always recognize the creaking of his boots. The door opened, and a figure appeared that was completely unknown to me.

\chapter{The Holy Fool} %c5

A man of about fifty entered the room; he had a pale, pockmarked, elongated face, long, grey hair, and a sparse, reddish beard. He was so tall that, in order to pass through the door, he not only had to bend his head down but bend at the waist as well. He wore something tattered that looked like a kaftan worn over a cassock; in his hand, he held an enormous staff.\footnote{The Russian kaftan was a coat or gown with long sleeves and a high collar, which buttoned down the front, very different from the garments that go by the name in other countries. --- Trans.} Entering the room, he knocked on the floor as hard as he could with it, and, his brows crooked and his mouth open outrageously wide, he roared with the most frightening and unnatural laughter. One of his eyes was crooked, and the white pupil of that eye danced incessantly and gave his already unattractive face an even more disgusting \todo{?} appearance.

``Aha! You're caught!'' he cried, running up to Volodya with short little steps, and grabbed him by the head, thoroughly inspecting it---then, with a perfectly serious expression he stepped back, went up to the table, and began to blow under the oilcloth and make the sign of the cross. ``Oh, misery! Oh, pain\ldots{}! My dearest ones\ldots{} they're flying away,'' he said, his voice trembling from tears, looking passionately and intently at Volodya, and started to wipe away his tears, which were indeed falling, with his sleeve. %grisha

His voice was hoarse and rough, his movements hurried and erratic, his speech nonsensical and rambling (he never used pronouns), but the cadence was so affecting, and his yellow, deformed face sometimes took on such an open, sorrowful expression, that, listening to him, it was impossible not to feel a mixture of pity, fear, and grief.

This was Grisha, a holy fool and pilgrim.

Where was he from? Who were his parents? What compelled him to choose the pilgrim's life that he led? No one knew. All I know is that since 1815 \todo{?} he was known as a holy fool, who winter and summer went around barefoot, visiting monasteries, giving little images of saints \todo{better than icons?} to those he came to love, and saying mysterious things, which certain people took as prophecies, that no one had ever known him in any other condition, that every now and then he would go to see grandmother, and that some said he was the unhappy son of rich parents and was pure of spirit, while others said he was just a peasant and an idler.

Finally, the long awaited and punctual Foka appeared, and we went downstairs. Grisha, sobbing and continuing to speak all kinds of absurdities, came after us and knocking his crutch against the steps of the stairs. Papa and maman were walking around the drawing room arm-in-arm and talking quietly about something. Marya Ivanovna sat sedately on one of the chairs that were arranged symmetrically at right angles to the sofa, and, with a strict but restrained voice gave instruction to the girls, who were next to her. \todo{look over that sentence again} As soon as Karl Ivanych entered the room, she looked at him and immediately turned away, and her face took on an expression that can be described as saying: I do not notice you, Karl Ivanych. By the girls' eyes it was obvious that they wanted badly to give us some very important piece of news; but to jump up from their places and go over to us would have been a violation of Mimi's rules. We first had to go over to her, say, ``Bonjour, Mimi!'' \todo{punct.} shuffle our feet, and only then were we allowed to enter into conversations.

What an intolerable individual that Mimi was! It always seemed that no topic was permissible around her: she found everything indecent. On top of that, she was forever nagging us: \textit{``parlez donc fran\c cais,''} and it was our bad luck that dinner was precisely where we wanted to chatter away in Russian; and just as you were digging into a dish and wanted to be left alone, then, without fail, it would be, \textit{``mangez donc avec du pain,''} or, \textit{``comment ce que vous tenez votre fourchette?''}\footnote{Speak French \ldots{} eat with bread \ldots{} how are you holding your fork? \textit{Fr.}} \todo{Added graf break here, not in original.} %mimi

``What business is it of hers?'' I would think. ``Let her teach her girls, we have Karl Ivanych for that.'' I shared his hatred for \emph{certain people} completely.

``Ask mama if you can take us along on the hunt,'' said Katyenka in a whisper, pulling on my jacket to stop me as the grown-ups were going ahead into the dining room. %katyenka

``Alright, we'll try.'' %nikolai

Grisha ate in the dining room, but at a special little table; he did not raise his eyes from his plate, sighed from time to time, made frightening faces, and said, as if talking to himself, ``A pity! She flew away\ldots{} The dove will fly away into the heavens\ldots{} oh, the stone on the grave!'' and so on. %grisha

Maman had been out of sorts since morning; Grisha's words and actions noticeably deepened her sad disposition. 

``Oh, yes, I almost forgot to ask you about one thing,'' she said, handing father a dish filled with soup. %maman

``What was it?'' %papa

``Tell them to lock up your frightening dogs, please; they very nearly bit poor Grisha when he came through the yard. They could attack the children, too.'' %maman

Hearing that they were talking about him, Grisha turned to the table, started showing everyone the tattered ends of his clothes, and said, while chewing:

``Wanted them to tear me to pieces\ldots{} God would not allow it. A sin to let the dogs torment me! A great sin! Don't beat the big man,\footnote{He called all men this, without distinction. --- \textit{LNT.}} what's a beating? God forgives\ldots{} Not in these days.'' %grisha

``What is he saying?'' asked papa, examining him with a fixed and severe gaze. ``I don't understand anything.'' %papa

``I understand him,'' answered maman: ``He told me that some hunter set his dogs on him on purpose, so he's saying, `He wanted them to tear me to pieces, but God would not allow it, and asks you not to punish him for it.''' %maman

``Ah! So that's it!'' said papa. ``How does he know that I want to punish that hunter?'' He then continued in French: ``You know that I am not very enthusiastic about these holy men, but that one especially bothers me, and he ought to\ldots{}'' %papa

``Ah, don't say that, my friend,'' maman interrupted him, as if frightened by something, ``what could you know about him?'' %maman

``Well, I suppose I've had the chance to study their sort of people---so many of them come to you---they're all alike. Always the same story\ldots{}'' %papa

It was clear that mother had an entirely different opinion about that and did not want to argue.

``Pass me that pie, please,'' she said. ``Well, are things going well today?'' %maman

``No, it makes me angry,'' papa continued, picking up the pie, but holding it at such a distance that maman could not reach it, ``no, it makes me angry when I see intelligent and educated people being taken in by lies.'' %papa

And he struck the table with his fork.

``I asked you to pass me the pie,'' she repeated, reaching out her hand.

``And it is wonderful,'' continued papa, moving his hand away, ``that the people like that get picked up by the police. The only good they do is upset the weak nerves of certain individuals,'' he added with a smile, noting that the conversation was bothering mother a great deal, and gave her the pie.

``I'll say only this: it's difficult to believe that a person who, in spite of his sixty years, walks around winter and summer barefoot and wears seventy pounds of chains under his clothing, never taking them off, and who more than once has turned down offers to live peacefully with all his needs met---it's difficult to believe that a person would do all of that just out of laziness.'' %maman

``Concerning the prophecies,'' she added with a sigh, falling silent for a moment, ``\textit{je suis pay\'ee pour y croire};\footnote{I have paid to be believed. \textit{Fr.}} \todo{check this French translation.} I'm sure I've told you how Kiryusha prophesied my late papa's death to the day and the hour.'' %maman

``Ah, what have you done to me!'' said papa, smiling and placing his hand by his mouth on the side where Mimi was sitting. (When he did this, I always listened with rapt attention, expecting something funny.) ``Why did you remind me of his feet? I just looked at them and now I won't eat anything.'' %papa

Dinner was coming to an end. Lyubochka and Katyenka were winking at us incessantly, turning in their chairs, and generally expressing serious agitation. Their winking meant: ``why aren't you asking if you can take us along on the hunt?'' I jabbed Volodya with my elbow, Volodya jabbed back, and finally we decided: first with a timid voice, then fairly firmly and loudly, he explained that, since we were going to leave that day, we would like it if the girls could go on the hunt with us in the brake. \todo{anachronism? or do I need to say shooting brake?} After a short conference among the grown-ups, the question was decided in our favor and---what was all the more pleasant---maman said that she herself would come with us.

\chapter{Preparation for the Hunt} %c6

During dessert, Yakov was called in and ordered to bring up the brake, the dogs, and the saddle horses---all in the greatest detail, calling each horse by name. Volodya's horse had a limp; papa gave orders for a hunter to be saddled for him. That word, ``hunter horse,'' sounded strange in maman's ears: it seemed to her that a hunter horse must be some kind of wild beast, and that it would certainly bolt and kill Volodya. Despite assurances from papa and Volodya, who with astonishing bravado said that it was nothing to him and that he loved it when a horse bolted, poor maman repeated over and over that she would be suffering the entire trip.

Dinner came to an end; the grown-ups went into the study to drink coffee, and we ran into the garden, shuffling our feet on the paths, which were covered with fallen, yellow leaves, and talking. We began conversations about the fact that Volodya would be riding a hunter horse, about how it was shameful that Lyubochka ran more slowly than Katyenka, about how interesting it would be to see Grisha's chains, and so on; about the fact that we would soon be parting not a word was said. Our conversation was interrupted by the clatter of the approaching brake, a house boy sitting on every spring. Behind the brake came the hunters with the dogs, and behind the hunters, the coachman, Ignat, on the horse designated for Volodya, leading my ancient nag. At first we rush to the fence, where we could see all of these interesting things, then with a shriek and a rush of feet, we ran upstairs to dress, and in fact to dress as much as possible like the hunters. One of the most important means for doing this was to stuff our trousers into our boots. Without the slightest delay, we got to work on this, hurrying to get done with it and run out to the front steps to revel in the sight of the dogs, the horses, and conversation with the hunters.

The day was hot. Strangely shaped white clouds had been gathering on the horizon since morning; then a little breeze began to drive them closer and closer, so that they covered the sun from time to time. No matter how close the clouds came or how dark they got, they were clearly not fated to become a storm and ruin our pleasure one last time. Toward evening, they started to break apart again: some turned white, elongated, and ran out to the horizon; others, just above our heads, turned into translucent white fluff; a single large black cloud remaind in the east. Karl Ivanych always knew where clouds like that would go; he explained that that cloud would go to Maslovka, that there would be no rain, and that the weather would be fine.

Foka, despite his advancing years, ran down the stairs very deftly and quickly, shouting, ``Let's go!'' and, setting his legs apart, stood firmly in the center of the porch, right between the spot where the coachman was supposed to bring up the brake and the threshold, positioned like a man who did not need to be reminded of his duty. The ladies came down and, after a short debate over who would sit on which side and hold on to whom (although it seemed to me there was no need to hold on to anyone), they found seats, opened their parasols, and were off. When the brake started moving, maman, pointing at the ``hunter horse,'' asked the coachman in a quivering voice:

``Is that the horse for Vladimir Petrovich?'' %maman

And when the coachman answered in the affirmative, she waved her hand dismissively and turned away. I felt extremely impatient: I mounted my horse, looked between its ears, and performed various evolutions in the yard.

``If you would be so kind, don't crush the dogs,'' one of the hunters said to me. %hunter

``Calm yourself: it's not my first time,'' I answered proudly. %nikolai

Volodya sat on the ``hunter horse,'' not without a small shudder, despite his strength of character, and, stroking its head, he asked several times:

``Is it gentle?'' %volodya

He looked very fine on a horse---just like a grown-up. His thighs sat so well in the saddle that I was jealous---especially because, as far as I could judge by my shadow, I looked less splendid by far.

Then we heard papa's steps on the stairs; the whipper-in drove together the hounds, which had scattered; the hunters with their borzois called theirs in and started to mount. The huntsman guided his horse up to the steps; the dogs from papa's pack, which had previously lain in various picturesque poses near the horse, ran toward him. After him came Milka, wearing a beaded collar and shaking her bell. She ran happily out as always to greet the kennel dogs: she played with one group of them, searched some for fleas, and smelled and growled at the others.

Papa mounted his horse, and we rode off.

\chapter{The Hunt} %c7

The whip, who was called the Turk, \todo{or the Ibrik? strange} rode in front of the rest on a blue roan with a hooked nose. He was wearing a fur hat and had an enormous horn slung over his shoulder and a knife on his belt. From this person's dark, grim appearance, one could easily think that he was on his way to a battle to the death, not a hunt. Near the hind legs of his horse, in a motley, excited tangle, ran the leashed hounds. A pitiful fate awaited any dog that decided to fall back. It required great effort to pull back its leash-mate, and when it had succeeded, one of the whippers-in riding behind would certainly slap her with the quirt, scolding her: ``back into the pack!'' Riding out through the gates, papa ordered the hunters and us to ride along the road while he himself turned into the rye field.

The grain harvest was in full swing. The endless, bright yellow field was enclosed on only one side by a tall, deep-blue forest, which at that moment seemed to me a far-off, mystical place, beyond which the earth must end, or some undiscovered lands begin. In the tall, thick rye, the bent back of a reaper woman could be seen in one of the cleared strips, the ears of grain flapping as she moved them back and forth between her fingers, another woman bent over a cradle, the sheaves scattered all over the stubble alongside the cornflowers. On the other side, men were standing in carts wearing only their long shirts, stacking the sheaves and shaking their dust onto the scorching-hot field. The foreman, in boots and a coat thrown over his shoulders, a tally stick in his hand, noticed papa from afar, took off his felt hat, wiped his red hair and beard with a towel, and shouted at the women. The chestnut horse that papa was riding trotted playfully, from time to time dropping its head to its chest to tug at the reins, and using its tail to brush away the flies that were hungrily clinging to it. Two borzoi dogs, curling their tails raptly like sickles and stepping high, jumped gracefully through the tall stubble behind the horses' feet; Milka ran forward and, curling up her head, waited for a treat. The hum of everyone's voices, the clatter of the horses and the carts, the happy singing of the quail, the buzzing of the insects that crowded the air, unmoving, the smell of the wormwood, \todo{?} straw, and horse sweat, a thousand different hues and shadows that were bathed in the light of the burning sun across the light yellow stubble, the deep blue expanse of the forest and the pale violet clouds, the white spider webs that floated in the air or lay on the stubble---all of this I saw, heard, and felt.

Approaching the Kalinovo Forest, \todo{?} we found the brake already there and, beyond all expectation, a one-horse cart as well, in the middle of which sat a server. Underneath the straw, we could see a samovar, a tub with a mold for making ice cream, and still more enticing little bundles and boxes. There was no mistaking it: it was tea in the open air, ice cream, and fruits. At the sight of the cart, we noisily expressed our happiness, because drinking tea in the forest on the grass, in a place where no one had ever, ever drunk tea, seemed like the greatest delight.

The Turk approached the island of grass, stopped, listened attentively to papa's lengthy instructions about how to divide the group and where to go (however, he never held to those instructions, but did as he pleased), unleashed the dogs, slowly tied the leashes to his saddle, mounted the house, and, whistling, disappeared into the young birches. The unleashed hounds first expressed their pleasure with a wag of their tails, shook themselves, straightened, and then with a slow little trot, sniffing and wagging their tails, ran off in different directions.

``Do you have a handkerchief?'' asked papa. %papa

I took mine out of my pocket and showed it to him.

``Well, tie your handkerchief on that grey dog\ldots{}'' %papa

``Zhiran?'' I said, with a knowing air. %nikolai

``Yes; and go run along the road. When you get to the clearing, stop and look: don't come back to me without a hare!'' %papa

I wrapped my handkerchief around Zhiran's furry neck and threw myself headlong toward my assigned place. Papa laughed and shouted after me:

``Quickly, quickly, or you'll be late.'' %papa

Zhiran kept stopping incessantly, raising his ears and listening to the hunters' calls. I did not have the strength to draw him from his place, and I began to yell: ``Hark! Hark!'' Then Zhiran burst forward with such strength that I could barely hold onto him and fell more than once before we reached our place. Choosing a shady, level place by the roots of a tall oak, I lay down on the grass, sat Zhiran down next to me, and started to wait. My imagination, as often happens on these occasions, ran far ahead of reality: I imagined myself chasing down my third hare while the first hound was just howling in the forest. Turka's voice rang out, loud and animated, throughout the forest; the hound yelped, and her voice was heard more and more often; a different, deeper voice joined it, then a third, a fourth\ldots{} The voices broke off, then broke in on top of one another. The sounds grew progressively stronger and more persistent and finally blended together into one ringing, rumbling din.  \textit{The island was clamorous, and the hounds were boiling over.}\footnote{The ``island'' refers to the unleashed pack of hounds chasing after the hare. --- Trans.}

Hearing this, I froze in place. Fixing my eyes on the edge of the forest, I grinned foolishly; sweat poured off of me, and although the drops, running down my chin, tickled me, I did not wipe them off. It seemed to me that no moment could be more decisive than this one. This attitude of intensity was too unnatural to continue for very long. The hounds rumbled along the very edge of the forest, then gradually moved away from me; there was no hare. I began to look around. With Zhiran, it was the same story: at first he was yelping and straining to be let loose, then he lay down beside me, put his muzzle on my knees, and calmed down.

Near the exposed roots of the oak under which I was sitting, along the dry, grey earth, among the dry, fallen oak leaves, the acorns, the shrunken, moss-covered twigs, the yellow-green moss, and the sparse, thin, green grasses, moved swarms upon swarms of ants. One after another, they hurried along the well-worn paths they had blazed: some of them with heavy loads, others traveling light. I picked up a long stick and blocked their way. You should have seen how some of them, not heeding the danger, crawled underneath, others climbed over; and a few, especially those who were carrying loads, were completely lost and did not know what to do: they kept stopping, looking for a way around, or they turned back, or they climbed up the stick to my hand and, it seemed, planned to climb under the sleeve of my jacket. I was distracted from these observations by a butterfly, with yellow wings, which hovered alluringly in front of me. As soon as I turned my attention toward it, it flew about two steps away from me, hovered over a wilted white clover blossom and landed on it. I do not know whether it was being warmed by the sun, or if it was taking nectar from that blossom---I only knew that it seemed to feel very fine. From time to time, it flapped its wings and pressed closer to the blossom, then finally it froze entirely. I laid my head on both hands and watched it with pleasure.

Suddenly, Zhiran howled and took off with such strength that I nearly fell. I looked around. At the edge of the forest, with one ear raised and the other pressed against its head, ran a hare. The blood rushed to my head, and I forgot everything at that moment: I shouted something in a furious voice, released the dog, and started to run. But no sooner had I done that than I began to regret it: the hare hunkered down, hopped once, and then it disappeared.

But I was even more ashamed when, chasing after the hounds and driving them on at the top of his voice, Turka appeared from behind the bushes! He saw my mistake (which lay in not \emph{waiting for the right moment}) and, glanced at me with contempt, said only: ``Ah, sir!'' But you had to hear how it was said! It would have been easier for me if he had hung me on the horn of his saddle like a hare.

I stood on that spot for a long time, despairing and not calling back the dog, just saying to myself as I slapped my thighs:

``My God, what did I do!'' %nikolai

I heard the hounds chasing further along, and the sounds of the island on the other side pursuing the hare, and Turka with his enormous horn calling the dogs---but I did not move from that spot\ldots{}

\chapter{Games} %c8

The hunt came to an end. In the shadows of the young birch trees, we spread a rug, and the whole group sat in a circle on the rug. The server, Gavrilo, having trampled the lush, green grass around him, wiped down the plates and took leaf-wrapped plums and pears out of a box. The sun shone through the green branches of the young birches and threw dancing circles of light onto the patterns of the rug, on my legs, and even on Gavrilo's bald, sweaty head. A light breeze, passing through the leaves of the trees, my hair, and my sweaty face, refreshed me considerably.

Once they gave us our ice cream and fruits, there was nothing to do on the rug, and we got up and went to play, in spite of the slanting, burning rays of the sun.

``Well, what game?'' said Lyubochka, squinting from the sun and hopping on the grass. ``Let's play Robinson.'' %lyubochka

``No, that's boring,'' said Volodya, dropping lazily to the grass and chewing on some leaves. ``You always want to play Robinson! If you want to play something, let's build a gazebo instead.'' %volodya

Volodya was clearly puffing himself up: he was obviously so proud of himself for riding on a hunter horse and was pretending to be very tired. It may also have been that he had too much reason and too little power of imagination to truly enjoy playing Robinson. The game consisted of acting out scenes from \textit{Robinson Suisse}, which we had all read not long before.\footnote{That is, \textit{The Swiss Family Robinson} by Johann David Wyss. --- Trans.}

``Oh, please\ldots{} Why don't you want to something to make us happy?'' the girls badgered him. ``You can be \textit{Charles} or \textit{Ernest}, or Father --- whichever one you want?'' said Katyenka, trying to pull him up from the ground by the sleeve of his jacket. %katyenka

``I really don't want to---it's boring!'' said Volodya, stretching out with a self-satisfied smile on his face.

``I wish we were sitting at home then, if no one wants to play,'' Lyubochka said through tears. %lyubochka

She was an awful crybaby.

``Well, let's go; just don't cry, please: I can't stand it!'' %volodya

Volodya's condescension gave us very little pleasure; on the contrary, his lazy and bored attitude took all the charm out of the game. When we sat on the ground and, pretending we were going fishing, began to row as hard as we could, Volodya sat with his arms crossed in a pose that had nothing in common with a fisherman's pose. I pointed this out to him; but he answered that, no matter how much or how little we waved our arms, we would not gain or lose anything and anyhow would not go far. Unfortunately, I had to agree with him. When, pretending that I was going hunting, with a walking stick over my shoulder, I set off for the forest, Volodya lay on his back, threw his hands behind his head, and told me that he was walking, too. Actions and words like that threw cold water on our games and were extremely unpleasant, all the more so because in our hearts we could not but agree that Volodya was acting reasonably.

Even I knew I could not kill a bird with a walking stick, or even shoot at one. It was a game. By that reasoning, you could not ride in a coach made of chairs; but Volodya would remember, I think, how on long winter evenings we used to cover an armchair with scarves, making a carriage out of it, with one of us the driver, the other the footman, the girls in the middle, and three chairs serving as the three horses---and we would go off down the road. What adventures we used to have on that road! And how happily and quickly those winter evenings went by! Judging by reality, no games are possible. And if no games are possible, then what is left?

\chapter{First Love, or Something Like It} %c9

Pretending that she was picking some kind of American fruits from the tree, Lyubochka plucked a worm of enormous size off one of the leaves, dropped it onto the ground with horror, raised her arms up and leapt back, as if afraid something would gush out of it. The game stopped; all of us, heads together, fell on the ground to look at this rare sight.

I looked over Katyenka's shoulder while she tried to lift the worm up on a leaf, which she had placed in its way.

I noticed that lots of girls had the habit of shrugging their shoulders, trying to get the open neck of their dress back into place with the motion. I also remember how Mimi would always get angry at that motion and say: \textit{``c'est un geste de femme de chambre.''}\footnote{That is a chamber maid's gesture. \textit{Fr.}} Bent over the worm, Katyenka made that very same motion, and at the same time the wind lifted her scarf from her little white neck. At the time she made that motion, her shoulder was two fingers away from my lips. I was not looking at the worm any more, I was looking as hard as I could at Katyenka's shoulder, and I kissed it. She did not turn, but I noticed that her neck and ears went red. Volodya, not even lifting his head, said contemptuously:

``Oh, what tenderness!'' %volodya

Tears came to my eyes.

I could not take my eyes off of Katyenka. I had long since gotten used to seeing her fresh little face and blond hair, and I had always loved them; but now I looked at her face more carefully and fell even more in love. When we came back to the grown-ups, papa, to my great joy, announced that, because maman had asked, our departure would be put off until the next day.

We rode back together in the brake. Volodya and I, trying to exceed one another in bravado and horseriding skill, kept prancing next to it. My shadow was longer than before and, judging by it alone, I concluded that I looked the part of an attractive horseman; but this self-satisfied feeling was soon demolished by the following circumstance. Trying to charm everyone sitting in the brake once and for all, I fell back a bit, then, with the help of the whip and my foot, I spurred on my horse, took on an easy, graceful attitude, and tried to rush past them like a whirlwind, on the side where Katyenka was sitting. I just had to decide which was better: to gallop by silently, or to shout something? But my intolerable horse, coming even with the draught horses, despite my every effort, came to a stop so unexpectedly that I came out of my saddle, hitting the horse's neck and nearly flying off of my mount.

\chapter{What Kind of Person Was My Father?} %c10

He was a person of the previous century and had in common with the youth of that century an elusive character that was chivalrous, intrepid, self-assured, courteous, and dissolute. He looked with contempt on people of the present century, and this view came as much from inborn pride as from a secret annoyance that he had not been able to have in our century the influence or the successes that he had had in his own. His two great passions in life we cards and women; he won in the course of his life several millions and had affairs with a countless number of women of all classes.

His large, stately stature, strange walk (tiny little steps), habit of twitching his shoulder, little eyes, always smiling, large, aquiline nose, odd lips, which fit together awkwardly but were pleasant-looking, deficiencies in pronunciation (a lisp), and a bald spot covering nearly his entire head: this was my father's outer appearance, which he knew how to use not only to pass as, and be, a person \textit{\`a bonnes fortunes,} but to please everyone without exception---people of all classes and stations, especially those he particularly wanted to please.

He knew how to take the upper hand in relations with anyone. Never a person of \emph{the very highest society}, he always associated with people of that circle, and was respected by them. He knew that utmost measure of pride and conceit which, without offending others, elevated him in the opinion of society. He was original, but not always, and employed his originality as a means of substituting in various circumstances for breeding or wealth. Nothing on earth could excite any feeling of surprise in him: whatever brilliant situation he found himself in, it seemed that he was born for it. He knew so well how to hide from others and keep at a distance from himself that dark side of life, so well known to all of us, filled with minor annoyances and troubles, that it was difficult not to envy him. He was a connoisseur of all things that give comfort and delight, and knew how to use them. Brilliant connections were his hobby, and he had them in part through my mother's relations, and in part through the comrades of his youth, with whom he was angry for going far up the ranks, while he always remained a retired lieutenant in the Guards. He, like all former military men, did not know how to dress according to fashion; he did, however, dress with originality and refinement. His clothes were always very wide and free, with splendid shirts, large, turned-back cuffs and collars\ldots{} However, it all fit his large stature, strong build, bald head, and calm, self-assured movements. He was sensitive and even prone to crying. Often, when reading aloud, when he came to a passage filled with pathos, his voice would begin to tremble, tears would start to show, and he would drop the book, annoyed. He loved music, and sang various songs, accompanying himself on the piano: romances by his friend A---, gypsy songs, and motifs from operas; but he did not love serious music and, paying no attention to general opinion, would say openly that Beethoven's sonatas aroused boredom and sleep in him, and that he liked nothing better than to hear ``Do not wake me, young lad'' as sung by Semyonova or ``Not alone'' as sung by the gypsy Tanyusha. His nature was one of those that needed a public for a good cause. \todo{?} And he only considered good what the public called good. God only knows whether he had any moral convictions whatsoever. His life was so full of passions of every kind that he had no time to form them, and he was so happy in life that he saw no need of them.

In old age he formed a permanent view on things and unchangeable rules---but solely on a practical basis: those actions and that form of life that provided him happiness or pleasure, he considered good, and found that always and everywhere, people should do exactly those things. He captivated people when he spoke, and that ability, I think, strengthened the flexibility of his rules: he was capable of describing one and the same action as either the dearest little prank, or the lowest dirty trick.

\chapter{Business in the Study and the Drawing Room} %c11

Dusk had already fallen when we arrived home. Maman sat down at the piano and we, the children, brought paper, pencils, paints, and settled in to draw all around the round table. I had only dark-blue paint; but, despite this, I decided to draw the hunt. Having made lively depictions of a blue boy riding a blue horse, and blue dogs, I felt uncertain whether I could draw a blue hare, and I ran to papa in his study to consult him about this. Papa was reading something and answered my question, ``Is there such a thing as a blue hare?'' without raising his head: ``there is, my friend, there is.'' Returning to the round table, I made my depiction of a blue hare, then found it necessary to turn the blue hare into a bush. I did not like the bush, either: I made it into a tree, the tree a haystack, the haystack a cloud, and finally I had so stained the paper with blue paint that I tore it up in annoyance and went to doze in the Voltaire armchair.

Maman played Field's second concerto---he had been her teacher. I dozed, and in my imagination sprang up bright memories, light and limpid. She played a Beethoven sonata, full of pathos, and I experienced memories of something sad, heavy, and dark. Maman played these two pieces often; and so, I remember well the feeling they awoke in me. That feeling was like a memory; but a memory of what? It seemed like you were remembering something that had never been.

 Across from me was the door into the study, and I saw Yakov enter along with some people in kaftans and beards. The door was closed immediately after they went in. ``Well, the business has started!'' I thought. It seemed to me that there must be no greater matters in the world that what transpired in that study; I was confirmed in this thought by the fact that everyone usually approached the doors of the study whispering and on tiptoe; papa's loud voice and the smell of his cigar came through, which always, I do not know why, attracted me greatly. Half-asleep, I was suddenly struck by the loud creak of boots in the footman's pantry. Karl Ivanych, on tiptoe, but with a dark and decisive face, with some kind of notes in his hand, approached the door and knocked lightly. He was let in, and the door again slammed shut.

``I hope nothing unfortunate will happen,'' I thought, ``Karl Ivanych is aggravated: he looks capable of anything\ldots{}'' %nikolai

I dozed off again.

Nothing unfortunate did happen; after an hour, I came to be awoken by the same creaking of boots. \todo{``came to be'' here to reflect the tone of ``cherez chas vremeni,'' possibly a better way out there to do that.} Karl Ivanych, wiping away tears, which I noticed on his cheeks, with his handkerchief, came out through the door and, mumbling something to himself under his breath, went upstairs. Papa came out after him and went into the drawing room.

``Do you know what I've just decided?'' he said in a happy voice, laying his hand on maman's shoulder. %papa

``What, my friend?'' %maman

``I'm going to take Karl Ivanych along with the children. There's room in the britzka.\footnote{A britzka was a long, hooded carriage, similar to a phaeton or chaise. Its greater length allowed for the inclusion of a bed or desk in the body of the carriage. --- Trans.} They are used to him, and he is quite attached to them, it seems; and 700 rubles a year is practically nothing, \textit{et puis au fond c'est un tr\`es bon diable.}''\footnote{And in essence he's a very good little devil. \textit{Fr.}} %papa

I could never get why papa called Karl Ivanych names.

``I am very glad,'' maman said, ``for the children, for him: he is a good old man.'' %maman

``If you had seen how touched he was when I told him to keep the 500 rubles as a gift\ldots{} but what is most amusing of all is the accounting he brought me. That is worth seeing,'' he added with a smile, giving her the note, written in Karl Ivanych's own hand, ``it's a delight!'' %papa

Here are the contents of that note:

% sic passim --- Karl Ivanych's broken written Russian
``For the childeren two of fishing rod --- 70 kopecks.

``Color paper, with a gold borders, plester, and a dummy for making boxes, in gifts --- 6 r.~55 k.

``Pant to Nikolai --- 4 rubles.

``Promised by Pyotr Aleksantrovich's from Moscow in the year 18--- a gold watch of 140 rubles.

``In sum it would result for Karl Mauer to receive besides his salaries --- 159 rubles 79 kopecks.'' %karl

Reading this note, in which Karl Ivanych demands to be paid all the money spent by him on gifts, and even to be paid for a promised gift, anyone would think that Karl Ivanych was nothing more than an insensitive and self-interested egoist---and anyone would be wrong.

Entering the study, notes in hand, and with a prepared speech in his head, he had planned to lay out eloquently before papa all the injustices he had suffered in our house; but when he began to speak in the same touching voice and with the same sensitive intonations with which he usually gave us dictation, his eloquence had its strongest effect on his own self; and so, arriving at the place where he said, ``however sad it would be for me to part ways with the children,'' he fell apart completely, his voice trembled, and he was forced to take checkered handkerchief out of his pocket. %karl

``Yes, Pyotr Aleksandrych,'' he said through tears (that part was definitely not in the prepared speech), ``I'm so used to the children that I don't know what I'll do without them. It should be better if I serve you without salary,'' he added, wiping away his tears with one hand and present his accounting with the other. %karl

That Karl Ivanych was speaking sincerely at that moment, I can affirm myself, because I know his kind heart; but how exactly his accounting accorded with his words remains a mystery to me.

``If you are sad, then it would make me even sadder to part ways with you,'' said papa, patting him on the shoulder, ``I've just changed my mind.''

Not long before supper, Grisha came into the room. From the very day he came into our house, he never ceased sighing and weeping, which, for many people who believed in his ability to prophesy, portended some kind of misfortune for our house. He started to say goodbye and informed us that he would move on tomorrow. I winked at Volodya and went through the door.

``What?'' %volodya

``If you want to see Grisha's chains, then let's go right now to the men's quarters upstairs---Grisha sleeps in the second room---we can sit in the storage room and see everything.'' %nikolai

``Excellent! Wait here, I'll call the girls.'' %volodya

The girls came running, and we went upstairs. Not without some argument over who would go into the dark storage room first, we found our seats and started to wait.

\chapter{Grisha} %c12

It felt eerie to us in the dark; we pressed close to one another and said nothing. Almost right behind us came Grisha, walking quietly. In one hand, he held his staff, in the other, a tallow candle in a brass candleholder. Not one of us breathed.

``Lord Jesus Christ! Most Holy Mother of God! Father, Son, and Holy Ghost\ldots{}'' he said over and over, breathing in air, with different intonations and abbreviations particular only to those who frequently repeat those words.

Setting his staff in the corner with a prayer and looking over the bed, he began to undress. Unfastening his ancient black belt, he slowly took off his tattered nankeen coat, smoothing it thoroughly and hanging it on the back of the chair. His face no longer expressed hastiness and stupidity; on the contrary, he was calm, pensive, and even majestic. His movements were slow and deliberate.

Remaining in just his shirt, he quietly lowered himself onto the bed, crossed himself from all sides, and he straightened the chains under his long shirt, obviously---because he winced---with some effort. Having sat for a little and looked attentively at his shirt, torn through in places, he stood up, lifted the candle up to the level of the icon case with a prayer, crossed himself before the several icons that hung there, and turned the flame of the candle upside-down. It went out with a snap.

Through the windows, which faced the forest, the nearly-full moon penetrated. The long, white figure of the holy fool was lit on one side by the pale, silver lunar rays, and on the other was a black shadow; together with the shadows from the window frames, it fell on the floor, walls, and reached to the ceiling. Out in the courtyard, the guard was knocking on his iron slab.

Having laid his enormous arms on his chest and lowered his head, his breathing incessant and hard, Grisha silently stood before the icons, then with difficulty lowered himself to his knees and began to pray.

At first, he quietly said well-known prayers, just stressing certain words, then repeating them, but louder and with great animation. He began to speak his words, with visible effort trying to express himself in Church Slavonic. His words were incoherent, but touching. He prayed for all his benefactors (that is what he called them when they took him in), among them for mother, for us, he prayed for himself, so that God would forgive him his grave sins, repeated: ``O God, forgive our enemies!'' then groaning, he got up, and, repeating over and over the same words, fell to the ground and again got up, despite the weight of his chains, which gave out a dry, sharp sound, striking the earth.

Volodya pinched my leg painfully; but I did not even look around: I merely ran my hand over the spot and continued, with a feeling of childish surprise, pity, and awe, to follow all Grisha's movements and words.

Instead of the fun and laughter that I had been counting on, entering the storage room, I trembled and felt my heart sink.

Grisha remained for a long time in that state of religious ecstacy and improvized prayers. First he would repeat a few times in a row: ``Lord, have mercy,'' but each time with new strength and feeling; then he would say, ``grant me your pardon, O Lord, teach me what to do\ldots{} teach me what to do, O Lord!'' with such feeling, that it was as if he expected an immediate answer to his words; then we would hear only mournful wailing\ldots{} He got up on his knees, laid his arms on his chest, and fell silent. %grisha

I slowly stuck my head out from behind the door and did not breathe. Grisha did not stir; heavy sighs escaped from his chest; in the cloudy pupil of his crooked eye, lit by the moon, stood a tear. 

``Thy will be done!'' he shouted suddenly with unsurpassable feeling, fell with his forehead on the earth, and wept like a baby. %grisha

Much water has flowed since that time, many memories of bygone days have lost their meaning for me and become vague dreams, and even the pilgrim Grisha has long since completed his final pilgrimage; but the impression that he made on me, and the feeling he awoke in me, will never die in my memory.

O great Christian, Grisha! Your faith was so strong that you felt the closeness of God, your love so great that the words flowed on their own from your lips---you did not subject them to reason\ldots{} And what praise you brought to His majesty when, not finding words, you fell to the earth in tears\ldots{}!

The feeling of deep emotion with which I listened to Grisha could not last long, because, first of all, my curiosity was more than satisfied, and second, my leg was worn out from sitting in one place and I wanted to join the whispering and racket of the group, which I heard behind me in the dark storage room. Someone grabbed me by the hand and said in a whisper, ``Whose hand is this?'' It was completely dark in the storage room; but by the contact and the whispering voice alone, right in my ear, I could tell right away it was Katyenka.

Absolutely unconsciously, I grabbed her arm, clad in short sleeves, by the elbow and pressed in toward her with my lips. Katyenka was truly surprised at my action and jerked back her hand: with that motion, she knocked over a broken chair that stood in the storage room. Grisha lifted his head, looked around quietly, and, saying a prayer, began to make the sign of the cross toward all corners of the room. We ran out of the storage room noisily, whispering to one another.

\chapter{Natalya Savishna} %c13

Halfway through the previous century, through the courtyards of the village of Khabarovka ran a barefooted but happy, fat, and red-cheeked girl in a shabby dress, \textit{Natashka}. Because of a request by her father, the clarinetist Savva, and long service, my grandfather brought her \emph{upstairs}---into the number of our grandmother's female servants. The lady's maid \textit{Natashka} distinguished herself in her position by her gentle disposition and her diligence. When my mother was born and needed a nanny, they entrusted that responsibility to \textit{Natashka}. In this new profession as well, she won praises and great rewards for her industry, loyalty, and affection for the young mistress. But the powdered head, stockings, and buckles of the young, sharp footman Foka, who in his service had frequent dealings with Natalya, captured her coarse but loving heart. She even decided to go herself to grandfather to ask for permission to marry Foka. Grandfather took her desire for ingratitude, became angry, and sent poor Natalya to the cattle yard in a village on the steppe as punishment. After 6 months, however, since no one could replace Natalya, she was returned to the courtyard and her previous position. Having returned from exile in her work clothes, she appeared to grandfather, fell before his feet, and begged him to return to her his favor and kindness and forget the foolishness that had come over her and which, she vowed, would never return. And indeed, she kept her word.

From that time, Natashka became Natalya Savishna and wore a matron's bonnet; the entire reserve of love she kept inside her, she transferred onto her young miss.

When a governess replaced her at my mother's side, she received the keys to the pantry, and the linens and all the provisions were given into her hands. She fulfilled these new duties with the same diligence and love. She lived her whole life for the good of the family, and saw waste, spoilage, and misuse everywhere, and used all her powers to try to fight the same.

When maman got married, desiring to show Natalya Savishna her thanks for twenty years of her labors and affection, she called her in and, expressing in the most complimentary terms all her appreciation and love for her, handed her a sheet of stamped paper on which was written Natalya Savishna's manumission, and said that, regardless of whether she would continue to serve in our house or not, she would always receive an annual pension of 300 rubles. Natalya Savishna silently listened to all of this, then, taking the document in her hands, glanced at it with spite, mumbled something through her teeth, and ran out of the room, slamming the door. Not understanding the reason for this strange action, maman, waiting a bit, went into Natalya Savishna's room. She was sitting with teary eyes on her trunk, working a handkerchief in her fingers, and was gazing fixedly at the scraps of her shredded manumission, which were scattered on the floor in front of her.

``What's with you, my dear, dear Natalya Savishna?'' asked maman, taking her by the hand. %maman

``Nothing, precious girl,'' she answered, ``I guess I've disgusted you somehow, since you're running me out of the house\ldots{} Well, I'll go.'' %natalya

She snatched away her hand and, barely holding back from crying, wanted to leave the room. Maman held her back, embraced her, and they both broke down crying.

As early as I can remember anything, I remember Natalya Savishna, her love and affection; but only now can I appreciate them---at the time, I did not have even a thought for how rare and marvelous a creature that old woman was. Not only did she not talk about herself, it seemed she did not think about herself: her entire life was love and self-sacrifice. I became so used to her selfless, tender love for us that I could not even imagine it could be any other way, I was not grateful to her in the least, and I never asked myself the questions: I wonder, is she happy? is she satisfied?

I used to run out of our lesson, under the pretext of unavoidable need, to her room, settle in, and start to daydream aloud, not in the least bothered by her presence. She was always busy with something: either she was mending a stocking, or rummaging in the trunks that filled her room, or taking inventory of the linens and, listening to whatever rubbish I was talking, ``so when I'm a general, I'm going to marry a marvelous beauty, buy myself a chestnut horse, build a glass house, and send for Karl Ivanych's relatives from Saxony,'' and so forth, she would say: ``yes, precious boy, yes.'' Usually, when I was getting up and preparing to leave, she would open a light blue trunk, inside the lid of which---as I now remember it---were glued: a painted representation of some kind of hussar, the picture from a jar of pomade, and one of Volodya's drawings---she would take a piece of incense out of that trunk, light it, and waving it around, would say:

``This is Ochakov incense, precious. When your late grandpa---God rest his soul---went to fight the Turk, he brought this back with him. This here is one of my last pieces,'' she would add with a sigh. %natalya

In the trunks that filled her room was absolutely everything. Whatever might be necessary, people usually said, ``You should ask Natalya Savishna,'' and, truly, after rummaging a bit, she would find the needed object and would say, ``Here, it's a good thing I saved this back.'' In these trunks were thousands of objects of the sort no one in the house except her worried about or knew. %natalya

One time I was angry with her. Here is what happened. At dinner, pouring myself some kvass, I dropped the carafe and spilled on the table-cloth.

``Go call Natalya Savishna, so she can enjoy what her little favorite just did,'' said maman.

Natalya Savishna came in, and, seeing the puddle that I had made, shook her head; then maman whispered something in her ear, and she, giving me a threatening look, left the room.

After dinner, I was going going into the hall, skipping, in a cheerful mood, when suddenly Natalya Savishna jumped out from behind the door with the table-cloth in her hand, grabbed me, and, regardless of desperate resistance on my part, started rubbing the wet spot in my face, saying, ``Don't you dare stain the table-cloth, don't you dare stain the table-cloth!'' This offended me so much that I started to howl from fury. %natalya

``How could she!'' I said to myself, walking around the hall and choking back tears, ``Natalya Savishna, just \emph{Natalya}, talking to me like that, ordering me around, and even hitting me in the face with that wet table-cloth like a servant boy. No, it's horrible!''\footnote{In Tolstoy's Russian text, Natalya Savishna commands Nikolai using the informal form of \textit{you} \textit{(ty)} rather than the formal \textit{(vy)}. Servants typically used the formal even with their masters' children. It's this informality that offends Nikolai in the original. --- Trans.}

When Natalya Savishna saw that I was starting to slobber, she immediately ran off, and I, continuing to walk around, thought about how I could pay back that impudent \emph{Natalya} for the abuse she had caused me.

After a few minutes, Natalya Savishna returned, came up to me timidly, and started to admonish me.

``Enough, my precious boy, don't cry\ldots{} Forgive me, I'm an old fool\ldots{} It's my fault\ldots{} Forgive me, now, dearie\ldots{} Here, take it.'' %natalya

She pulled out a cone out of her scarf, made of red paper, which had two caramels and a dried fig in it, and handed them to me with trembling hand. I did not have sufficient strength to look the kind old woman in the eye; I took the present, turning away, and my tears flowed even more abundantly, no longer out of fury now, but out of love and shame.

\chapter{Parting} %c14

On the next day, after the events I described, at twelve o'clock in the morning, a barouche and a britzka stood by the porch. Nikolai was dressed for the road, that is, his trousers were tucked into his boots and his belt was wrapped tight-tight around his old frock coat. He stood in the britzka and was putting overcoats and pillows away underneath the seating; when it seemed too high to him, he would sit down on the pillows and, hopping up and down, crammed them down.

``Do me a little favor, Nikolai Dmitrych, would you find a place for the master's little case,'' said papa's breathless valet, leaning out of the carriage, ``it's quite small\ldots{}'' %valet

``I wish you had said so earlier, Mikhei Ivanych,'' answered Nikolai rapidly and with annoyance, throwing some parcel onto the bottom of the britzka with all his might. ``Oh God, my head is spinning already, and here you come with your little cases,'' he added, lifting his peaked cap and wiping big drops of sweat from his tanned forehead. %nikolai

The servant men in their frock coats, kaftans, and shirts, without hats, the women, in their work clothes, striped scarves, with babies in their arms, and the barefooted kids all stood by the front steps, looking at the carriages and talking among themselves. One of the drivers---a hunched old man in a winter hat and a rough, hooded cloak---held the barouche pole in his hand, pushed it this way and that, and thought deeply and carefully about how the carriage would ride; the other---a conspicuous young fellow wearing only a long white shirt with Turkey red gussets, in a black felt hat that looked like a dinner plate, which he kept knocking first toward one ear, then the other, scratching his blond curls---laid his own cloak on the coach box, tossed the reins there as well, and, flipping a little, braided whip back and forth, looked first at his boots, then at the coachmen who were greasing the britzka. One of them, straining, held the carriage; \todo{close?} the other, bent over the wheel, thoroughly greased the axle and bushing---in fact, so that no leftover tar was wasted on the brush, he even dabbed some on the bottom, making circles. \todo{?} The worn-out post horses, of all different colors, stood near a railing and swatted at flies with their tails. Some of them, spreading their shaggy, swollen legs, squinted their eyes and dozed off; others, out of boredom, scratched one another or nibbled on leaves or the tough stalks of dark-green ferns that grew near the front steps. A few of the borzoi dogs---some breathed heavily, lying in the sun, others walked in the shade under the barouche and the britzka and licked the fat near the axles. Everywhere in the air was a kind of dusty haze, the horizon was a grey-violet color; but there was not a cloud in the sky. A strong western wind lifted columns of dust from the roads and paths, bent the tops of the tall linden trees and birches in the garden, and carried the fallen yellow leaves far away. I sat by the window and waited impatiently for the end of all the preparations.

When everyone had gathered in the drawing room around the round table, so that we could spend a few minutes together one last time, it never even crossed my mind that a sad moment awaited us. Completely empty thoughts ran through my mind. I asked myself various questions: which driver would go in the britzka and which one in the barouche? who would go with papa, and who with Karl Ivanych? and why did they want to make certain I was wrapped up in a scarf and a padded coat?

I thought, ``Am I a little baby? I'm sure I won't freeze. I just wish it would all be over: let's just get in and go.'' %nikolai

``Who should I give the note about the children's clothing to?'' said Natalya Savishna to maman, coming into the room with teary eyes and a note in her hand. %natalya

``Give it to Nikolai, then you come and say goodbye to the children.''

The old woman was about to say something, then suddenly stopped, covered her face with her scarf, and, waving her hand, left the room. My heart ached a bit, when I saw that movement; but my impatience to leave was stronger than that feeling, and I continued to listen with perfect indifference to father's conversation with mother. They were talking about things that obviously did not interest the one or the other: what needed to be bought for the house? what needed to be said to Princess Sophie and Madame Julie? and would the road be good?

Foka entered and, in precisely the same voice he used to announce, ``The food is ready,'' stopped by the lintel and said, ``The horses are ready.'' I noticed that maman winced and went white at the news, as if it were unexpected for her.

Foka was ordered to close all the doors in the room. This amused me greatly, ``as if we're all hiding from someone.''

When everyone was seated, Foka also sat down on the edge of a chair; but as soon as he had done that, the door squeaked and everyone turned to look. It was Natalya Savishna, hastily coming in, and, without raising her eyes, she took shelter near the door on the same chair as Foka. I can see before me now like it was yesterday Foka's bald head and wrinkled, immobile face, and the bent, kind figure in the bonnet, with grey hair visible underneath. They were pressed together on the same chair, and both of them were uneasy.

I continued to be carefree and impatient. The ten seconds that we spent sitting with the door closed seemed to be an entire hour. Finally, we stood, crossed ourselves, and started to say goodbye. Papa embraced maman and kissed her a few times.

``Enough, my dear,'' said papa, ``we're not parting forever, after all.'' %papa

``It's still sad!'' said maman with a voice quivering from tears. %maman

When I heard that voice, and saw her trembling lips and her eyes full of tears, I forgot about everything and was sad myself, pained and so afraid, that I would rather have run away than say goodbye to her. I understood in that moment that, while embracing my father, she had already said her goodbye to us.

She began to kiss Volodya and make the sign of the cross over him so many times, that---supposing she would turn to me next---I shoved myself forward; but she kept blessing him and pressing him to her chest over and over. Finally, I embraced her and, clinging to her, cried and cried, thinking of nothing else but my sorrow.

When we went to get in, the annoying servants came into the front room to say goodbye. Their ``your hand, please,'' noisy kisses on our shoulders, and the smell of lard from their heads awoke in me a feeling very close to distress at the presence of these irritating people. Under the influence of this feeling, I kissed Natalya Savishna's cap with extreme coldness when she, all in tears, said goodbye to me.

It is strange that I can see before me now all the faces of the servants and could describe them in the finest detail; but maman's face and posture absolutely escape the grasp of my imagination: possibly because during all that time I could not once get up the courage to look at her. It seemed to me that, had I done that, her and my sorrowfulness would grow to impossible heights.

I threw myself before everyone else into the barouche and settled on the rear seat. The top was up and blocked my view, but some kind of instinct told me that maman was still there.

``Should I look at her, or no\ldots{}? Oh, just one last time!'' I said to myself and leaned out of the barouche toward the front steps. At that time, maman, thinking the same thing, went up to the opposite side of the barouche and called me by name. Hearing her voice behind me, I turned toward her, but so quickly that we knocked heads; she smiled sadly and kissed me hard, hard for the last time.

When we had driven a few yards, I decided to look at her. The wind had picked up the light blue scarf that was tied around her head; her head lowered and her face covered with her hands, she was slowing climbing the front steps. Foka was supporting her.

Papa sat next to me and said nothing; I was choking back tears, and something was pinching my throat such that I was afraid I might suffocate\ldots{} Driving out onto the main road, we saw a white handkerchief, which someone was waving from the balcony. I continued to cry, and the thought that my tears showed my sensitive nature gave me pleasure and joy.

After we had gone a verst or so, I settled in more comfortably and began to look with persistent attention at the nearest object in front of my eyes---the rear part of the trace-horse, which was running on my side.\footnote{A verst was a unit of length roughly two-thirds of a mile (3,500 feet) long. ---Trans.} I watched how that trace-horse, a pinto, waved its tail, how it knocked one leg against the other, how the driver touched it with the braided whip and its legs started to leap together; I watched how the back band leapt on it, and the rings after the back band, and watched until the back band became covered in lather near the tail. I began to look around: at the wavy fields of ripe grain, at the dark fallows, where I could see a plow, a peasant, a horse with a foal, at the verst markers, I even looked at the coach box to see which driver was riding with us; and my face was not even dry from my tears before my thoughts were far from my mother, with whom I just had parted, perhaps forever. But every memory brought me back to thoughts of her. I remembered the mushroom that I had found the day before in the birch lane, remembered how Lyubochka and Katyenka had fought over who would get to pick it, remembered how they had cried when saying goodbye to us.

How I was sad to leave them! sad for Natalya Savishna, and the birch alley, and sad for Foka! Even that awful Mimi---I was sad to leave her. Sad, sad to leave everything! And poor maman? Tears welled up again in my eyes; but not for long.

\chapter{Childhood} %c15

O happy, happy, forever-lost days of childhood! How could I not love, not cherish my memories of that time? These memories revive and elevate my soul and serve as a source of the greatest enjoyment for me.

When tired of running about, at times I would sit at the tea table, on my high chair; it is late, and I have long since drunk my cup of milk with sugar, sleep is beginning to shut my eyes, but I do not move from my place, I sit and listen. How could I not listen? Maman is talking with someone, and the sounds of her voice are so sweet, so inviting. Those sounds alone say so much to my heart! With eyes dimmed with sleep, I look fixedly at her face and, suddenly, she becomes small, so small---her face is no larger than a button; but it is still clearly visible: I see how she glances at me and smiles. I love seeing her so tiny. I squint my eyes even further, and she becomes no larger than when I see myself, a boy, reflected in others' pupils; but then I stir---and the spell is destroyed; I narrow my eyes, turn around, try every which way to bring it back, but in vain.

I get up, get my legs together, and settle comfortably into my chair.

``You'll fall asleep again, Nikolenka,'' maman says to me, ``you'd better go upstairs.'' %maman

``I don't want to sleep, mama,'' I answer her, and obscure but sweet visions fill my imagination, healthy, childish sleep begins to shut my eyelids, and after a minute, I lose consciousness and sleep until someone wakes me. I would feel, half-waking, someone's tender hand shaking me; from the touch alone I could recognize it and, still asleep, involuntarily grab it and press it hard, hard to my lips. %nikolai

Everyone has already gone; one candle is burning in the drawing room; maman said she would wake me up herself; it is she who sits in the chair on which I am sleeping, her marvelous hand stroking my hair, and above my ear comes that lovely, familiar voice:

``Get up, dear soul: it's time to go to sleep.'' %maman

No one's indifferent gazes constrain her: she is not afraid to pour out all her tenderness and love on me. I do not stir, but kiss her hand all the harder.

``Get up, my angel.'' %maman

With her other hand, she takes me by the collar and, her fingers moving quickly, she tickles me. It is quiet in the room, half-dark: my nerves are excited by the tickling and from being woken up; mama sits right next to me; she touches me; I smell her scent and hear her voice. All of this makes me jump up, throw my arms around her neck, press my head to her chest, and, gasping, say:

``Oh, lovely, lovely mama, how I love you!'' %nikolai

She smiles her sad, charming smile, takes my head in both her hands, kisses my forehead, and sits me on her knees.

``So you love me very much, do you?'' She is silent for a minute, then says, ``see that you always love me, never forget me. If your mama is gone from this world, you won't forget her? You won't forget, Nikolenka?'' %maman

She kissed me even more tenderly.

``Enough! Let's not talk about this, my dearest mama, my soul!'' I cry, kissing her knees, and tears flow in streams from my eyes---tears of love and delight. %nikolai

After this, when I would arrive upstairs and kneel before the icons, in my cotton dressing-gown, what a marvelous feeling I would experience, saying: Lord, bring papa and mama to salvation. Repeating my prayers---which my mouth, in early childhood, mumbled for the first time before my beloved mother---love toward her and love toward God became strangely mixed in one feeling.

After my prayers, I would wrap myself up in the covers; my soul would feel weightless, bright, and comforted; one dream would come right after another---but what were they about? They were elusive, but filled with pure love and hopes for joy and happiness. I would remember Karl Ivanych and his bitter fate---the only person that I knew to be unhappy---and I would feel so miserable, and love him so much, that tears would flow from my eyes, and I would think, ``God grant him happiness, give me the opportunity to help him, to ease his sorrow; I'm ready to sacrifice anything for him.'' Then I would bury my favorite porcelain toy---a little bunny or doggy---in the corner of a down pillow and admire how nice, comfortable, and warm it was for it to lie there. I would also pray for God to give happiness to everyone, for everyone to be contented, and for there to be good weather for going out the next day, then I would turn onto my other side, my thoughts and dreams would get all mixed up, I would laugh, and fall asleep quietly, peacefully, with my face still wet with tears.

Will that freshness, that carefree feeling, the need for love and strength of faith I possessed in childhood ever return? What time could be better than when the two greatest virtues---innocent joy and a boundless need for love were the only impulses in life?

Where are those heated prayers? Where is that best of gifts---those pure tears of tender emotion? My guardian angel would fly down, wipe away those tears with a smile, and send down sweet visions to my unspoiled, childish imagination.

Has life really left such heavy tracks on my heart that those tears and delights are gone from me for ever and ever? Am I really left with memories alone?

\chapter{Verses} %c16

Almost a month after we moved to Moscow, I was sitting upstairs at my grandmother's house, at a big table, and was writing; opposite me sat the drawing teacher making some final corrections to a drawing in black pencil of a Turk's head wearing a turban. Volodya stood behind the teacher, craning his neck, and watched him from over his shoulder. The head was Volodya's first work in black pencil and that very day, on grandmother's angel day, he was supposed to bring it to her.\footnote{An angel day, also called the name day, was a person's baptismal anniversary. ``Angel'' references the character's grandmother's namesake, see note XX above. --- Trans.} \todo{Create reference to first footnote in ch. 1.}

``So you're not going to lay down shadows here?'' said Volodya to the teacher, standing on tip-toe and pointing at the Turk's neck. %volodya

``No, it's not necessary,'' said the teacher, putting away the pencils and drawing-pen in a box with a sliding lid, ``now it looks splendid. Don't lay a finger on it anymore. Well, and you, Nikolenka,'' he added, getting up and continuing to look sidelong at the Turk, ``will you finally reveal to us your secret? What are you bringing your grandmother? To tell the truth, you'd best bring a head, too. Goodbye, gentlemen,'' he said, took up his hat and ticket, and left.

In that moment, I also thought a head would be better than what I had been working on. When they told us that it would soon be grandmother's name day and that we should prepare a present for the day, the thought crossed my mind to write some verses for her for the occasion, and I put down two verses right away, with rhymes, hoping to put down the rest just as quickly. I absolutely cannot remember how I got that idea, so strange for a child, but I remember that I liked it a great deal, and that I answered every question on the subject by saying that I would certainly bring grandmother a present, but that I would not tell anyone what it was going to consist of.

Against all my expectations, it turned out that, besides the two verses I had thought up in the heat of the moment, I could not, despite all my efforts, compose anything further. I began to read the verses that were in our books; but neither Dmitriyev nor Derzhavin could help me;\footnote{Ivan Ivanovich Dmitriyev and Gavri\"il Romanovich Derzhavin were renowned poets and civil servants in the late 18th and early 19th centuries. --- Trans.} on the contrary, they further convinced me of my lack of ability. Knowing that Karl Ivanych loved to copy down little verses, I started to rummage around in his papers and, among some German poems, I found one, Russian, apparently from his own pen.

\begin{verse}
Madame L\ldots{}\ldots{} Petrovskaya, 1828, June 3.

Remember here,\\
Remember there,\\
Remember me ever.\\
From this today and to forever\\
Remember always to my grave\\
How I truly love to you gave.

Karl Mauer.
\end{verse}

I liked this poem, written in a pretty, round hand, on a thin piece of postal paper, for the touching feeling that permeated it; I learned it by heart at once and decided to take it as a pattern. Things went much more easily. On the name day, a congratulatory poem of twelve verses was ready, and, sitting at the table in the classroom, I copied it onto vellum.

Two sheets of paper were already ruined\ldots{} but not because I had thought to change anything in them: the verses seemed very fine to me; but from the third line, their ends began to turn more and more toward the top, so that from a distance it was obvious that they were written crookedly and were no good for anything.

The third sheet was just as crooked as the previous ones; but I decided not to copy anymore. In my poem, I congratulated grandmother, wished her many years of health, and concluded with this:

\begin{verse}
We'll love you this year and another,\\
And praise you like our own true mother.
\end{verse}

It seemed like it might not come out too badly, but this last verse offended my ear in some strange way.

``And praise you like-our fam-ily's mo-ther,'' I repeated under my breath. ``What rhyme could I use in place of \emph{mother}? Brother? Smother\ldots{}? Oh, it will work! It's better than Karl Ivanych's!'' %nikolai

So I wrote down the final verse. Then I read the whole work aloud in the bedroom, with feeling and gestures. There were lines that had no meter at all, but I did not stop for them; the last one struck me even more strongly and unpleasantly. I sat down on the bed to think\ldots{}

``Why did I write: \textit{like our own true mother}? Our actual mother was not there, after all, it was hardly necessary to mention her; I loved my grandmother, it was true, respected her, but all the same, she was not like my mother\ldots{} why had I written that, why had I lied? It \emph{was} poetry, but it still was not necessary.'' %nikolai

At the same time, the tailor came in carrying a new jacket.

``Well, so be it!'' I said with intense impatience, tossed the verses under my pillow with annoyance, and ran to try on my Moscow clothes. %nikolai

My Moscow clothes turned out to be very fine: a brown jacket with bronze buttons, tailored to fit tightly---nothing like the ones they made for us in the village, with room to grow---black pants, also narrow, and it was a marvel how well they brought out my muscles and broke over my boots.

``Finally I have pants with trouser stripes on them, real ones!'' I dreamed, beside myself with joy, looking at my legs from all angles. Although I felt uncomfortable, pressed tightly into the new clothes, I hid that from everyone and said that, on the contrary, I was very much at ease, and that if there was any shortcoming in the clothing, it was only that it was a little too roomy. After that, I stood for a long time in front of the mirror coming my abundantly pomaded hair; but no matter how I tried, I could not smooth the tufts that stuck up from the back of my head: as soon as I stopped pressing on them with the brush to test whether they would obey, they stood up and stuck out in different directions, making my face look very funny.

Karl Ivanych was getting dressed in the other room, and someone brought his blue tail coat and some other white articles of clothing to him through the classroom. At the door leading downstairs, I could hear the voice of one of grandmother's maids; I went out to see what she needed. She held a heavily starched shirt-front over her arm and told me that she was bringing it to Karl Ivanych, and that she had not slept the night before because she was getting it washed in time. I took the shirt-front and asked whether grandmother had gotten up.

``Come now! She has already had her coffee and the archpriest has arrived. What a handsome youngster you are!'' she added with a smile, looking over my new clothes. %maid

This remark made me blush; I turned around on one foot, snapped my fingers, and hopped, wanting to make her feel that she truly knew just how much of a handsome youngster I was.

When I brought the shirt-front to Karl Ivanych, it was already unnecessary: he had put on another and, bending over in front of a little mirror that stood on the table, he was holding the puffy bow of his tie with both hands and was trying to see whether his smoothly shaven chin would fit freely through it or not. Having adjusted our clothes from all sides and having asked Nikolai to do the same for him, he led us to grandmother. I find it funny to remember how strongly all three of us smelled of pomade while we began to go down the staircase.

Karl Ivanych had in his hands a little box of his own making, Volodya had his drawing, and I my verses; each of us had a greeting on the tip of our tongues to accompany our gift. At the moment when Karl Ivanych opened the door to the hall, the priest was putting on his \textit{riza}, and the first sounds of prayer rang out.\footnote{\textit{Riza} is one name for the upper vestment worn by a priest during an Orthodox worship service.}

Grandmother was already in the hall: hunched and leaning on the back of a chair, she stood by the wall in pious prayer; papa stood near her. He turned to us and smiled, noticing how we hurriedly hid the gifts that we had prepared behind our backs and, trying to remain unnoticed, stood right by the door. The whole effect of surprise that we were counting on was lost.

When everyone began to approach the cross, I suddenly felt myself under the heavy influence of an insurmountable, stupifying shyness and, feeling that I would never have enough courage to bring my gift, I hid behind Karl Ivanych's back; he greeted grandmother with carefully chosen expressions, passed his little box from his right hand to his left, delivered it to the \textit{honor\'ee}\todo{this is a bit dicey, no French in the original}, and walked back several steps to make way for Volodya. Grandmother seemed to be enraptured by the little box, which had gold borders pasted on it, and expressed her gratitude with the most tender smile. It was obvious, however, that she did not know where to put the little box, and that was probably why she offered it to papa to see how incredibly artfully it was made.

Having satisfied his curiosity, papa handed it to the archpriest, who seemed to greatly enjoy the little thing: he shook his head and looked with curiosity now at the little box, now at the master who could make such a splendid thing. Volodya brought up his Turk and also won the most flattering compliments from all sides. Then came my turn: grandmother turned to me with an approving smile.

Those who have experienced shyness know that it increases in direct relationship to time, while decisiveness decreases in inverse relationship, that is: the longer the condition continues, the more insurmountable it becomes and the less decisiveness remains.

The last of my bravery and decisiveness left me while Karl Ivanych and Volodya were bringing up their presents, and my shyness surpassed all bounds: I felt my blood rushing incessantly to my head as one blush gave way to another and big drops of sweat formed on my forehead and nose. My ears burned, I felt a shudder and perspiration over my entire body, and I shifted from one foot to the other and could not move from my place.

``Well, show me what you have, Nikolenka---a little box or a drawing?'' papa said to me. There was nothing to be done: with a trembling hand I held up the fateful sheet, all crumpled; but my voice completely refused to serve me, and I stopped in front of grandmother, silent. I was driven out of my mind by the thought that, instead of the expected drawing, in front of everyone they were going to read my good-for-nothing verses and the words: \textit{like our own true mother}, which would clearly show that I had never loved her and had forgotten her. How can I convey my sufferings when grandmother began to read my poem aloud and when she stopped, lost, in the middle of a verse to smile at papa in what seemed to me then to be a mocking way, when she spoke it not at all the way I wanted it to be read, and when, because of her weak eyesight, not reading to the end, she gave the paper to papa and asked him to read it to her from the beginning? It seemed to me that she did it because she was tired of reading such bad and crookedly written verses, and so that papa could himself read the last verse, which so clearly proved my callousness. I waited for him to smack me on the nose with the verses and say, ``Awful little boy, never forget your mother\ldots{} Here's something to think about!'' But nothing like that happened; on the contrary, when everything had been read, grandmother said: \textit{``charmant!''} and kissed me on the forehead. %papa

The little box, the drawing, and the verses were laid next to two cambric shawls and a snuff box with a portrait of maman on the pullout table of the Voltaire chair in which grandmother always sat.

``Princess Varvara Ilyinichna,'' announced one of the two enormous footmen who rode on grandmother's carriage. %footman

Grandmother was lost in thought, looking at the portrait that was set in the tortoise-shell snuff box, and did not answer.

``Should I admit her, Your Excellency?'' repeated the footman.

\chapter{Princess Kornakova} %c17

``Admit her,'' said grandmother, settling deeper in her chair.

The princess was a woman of about forty-five, small, frail, dried-up and peevish, with greyish-green, unpleasant eyes, whose expression clearly contradicted the unnaturally sweet set of her little mouth. Light red hair showed from beneath her velvet hat with its ostrich feather; her brows and eyelashes were even brighter and redder against the unhealthy color of her face. Despite this, thanks to her natural movements, tiny little hands, and her particularly dry features, her appearance overall had something noble and energetic about it.

The princess talked a great deal, and, because of her talkative nature, belonged to that category of people who always talk as if they are being contradicted, even though no one has said a word to them: she would raise her voice, then, gradually lowering it, she would suddenly begin talking with new liveliness and would look around at those individuals present who were not taking part in the conversation, as if trying to buck herself up with a look.\todo{ugh.}

Despite the fact that the princess kissed grandmother's hand and called her \textit{ma bonne tante} incessantly,\footnote{My good aunt. \textit{Fr.}} I noticed that grandmother was displeased with her: she raised her brows in a sort of unusual way, listening to her story about how it was impossible for Prince Mikhaylo to come and congratulate grandmother, despite the strongest possible desire; and, answering the princess's French speech in Russian, she said, stretching her words more than usual:

``I am very grateful to you, my dear, for your attentiveness: and as for Prince Mikhaylo not coming, well, what can one say\ldots{} he always has masses of things to attend to; and how could one say it would be a pleasure for him to sit with an old woman?'' %grandmother

And, not giving the princess time to rebut her words, she continued:

``So, my dear, how are your little children?'' %grandmother

``Yes, thank God, \textit{ma tante}, they're growing up, studying, playing little tricks\ldots{} especially Etienne, my oldest, he's becoming such a rascal that there's no peace and harmony in the house; though he is intelligent---\textit{un gar\c con, qui promet.}\footnote{A boy with promise. \textit{Fr.}} You can only imagine, \textit{mon cousin,}'' she continued, turning to papa alone, because grandmother, not the least interested in the princess's children but wishing to brag about her grandsons, took my verses with care from under the little box and began to unfold them: ``you can only imagine, \textit{mon cousin}, what he did the other day\ldots{}'' %kornakova

And the princess, leaning toward papa, began telling him the story with great animation. Having finished her story, which I did not hear, she began laughing at once and, looking searchingly into papa's face, she said:

``What do you think of the boy, \textit{mon cousin}?\todo{correct italics?} He deserved to be whipped; but the idea of the trick\todo{or is it the lie he told to cover?} was so intelligent and amusing that I forgave him, \textit{mon cousin.}'' %kornakova

And the princess, fixing her eyes on grandmother and saying nothing, continued to smile.

``Do you really \textit{beat} your children, my dear?'' asked grandmother, raising her eyebrows meaningfully and placing special emphasis on the word \textit{beat}. %grandmother

``Ah, \textit{ma bonne tante,}'' casting a quick glance at papa, the princess answered in a sweet little voice, ``I know your opinion on that count; but permit me not to agree with you on this one thing: however much I think about it, however much I read or take advice on this subject, experience still brings me back to the certainty that it is necessary to use fear on children. To make anything of a child, fear is necessary\ldots{} isn't that so, \textit{mon cousin}? And what, \textit{je vous demande un peu,}\footnote{I ask you. \textit{Fr.}} do children fear more than the rod?'' %kornakova

With this, she looked at us searchingly, and, I admit, I was somehow uncomfortable in that moment.

``Whatever you say, a boy is still a baby until 12 or even 14 years old; now a girl, that's another matter.'' %kornakova

``How fortunate,'' I thought, ``that I am not her son.'' %nikolai

``Yes, that is splendid, my dear,'' said grandmother, folding up my verses and laying them under the little box, as if after this she did not consider the princess worthy of hearing such a composition, ``that is very good, only tell me, please, just what kind of delicate feelings can you expect from your children after that?'' %grandmother

And, considering that argument irrefutable, grandmother added, to end the conversation:

``However, everyone is entitled to their opinion on that count.'' %grandmother

The princess did not answer, but only smiled condescendingly, expressing by this that she forgave these strange prejudices in an individual whom she esteemed so greatly.

``Ah, now introduce me to your young people,'' she said, looking at us and smiling invitingly. %kornakova

We stood up and, turning our eyes to the princess's face, we had no idea what we needed to do to show that we had been introduced.

``Kiss the princess's hand,'' said papa. %papa

``I beg you to love your old aunt,'' she said, kissing Volodya's hair, ``although I'm a distant relative, I count by the ties of friendship, not by degrees of relation,'' she added, directing this predominantly to grandmother; but grandmother continued to be displeased with her and answered:

``Eh! My dear, is that really what passes for family nowadays?'' %grandmother

``This is my young man of the world,'' said papa, pointing at Volodya, ``and that is the poet,'' he added while I was kissing the small, dry little hand of the princess and imagining with extreme clarity a rod in that hand, beneath the rod, a bench, and so on, and so on. %papa

``Which?'' asked the princess, holding me by the hand. %kornakova

``That one, the little one with the curls,'' answered papa, smiling cheerily.

``What did my curls ever do to him\ldots{}is there really nothing else for him to talk about?'' I thought and walked away to the corner.

I had the strangest ideas about beauty---I even considered Karl Ivanych the handsomest man in the world; but I knew very well that I was not good looking, and in that I was not mistaken one bit; and so every comment about my physical appearance offended me.

I remember very well a time when, at dinner---I was six years old then---they were talking about my physical appearance, and maman tried to find something good to say about my face: she said that I had intelligent eyes, a pleasant smile, and, finally, ceding the argument to my father and plain fact, was forced to admit that I was ugly; and then, when I thanked her for dinner, she patted me on the cheek and said:

``You remember this, Nikolenka, no one is ever going to love you for your face; so you have to try to be an intelligent and kind boy.'' %maman

Those words not only convinced me that I was not handsome, but that I should certainly be a kind and intelligent boy.

Despite this, I frequently found myself experiencing moments of despair: I imagined that there was no happiness on earth for a person with such a wide nose, fat lips, and small, grey eyes like mine; I asked God to perform a miracle---turn me into a handsome boy, and everything that I had at present, everything that I would have in the future, I would give it all for a handsome face.

\chapter{Prince Ivan Ivanych} %c18

Once the princess had heard the verses and heaped their composer with praises, grandmother softened, began to talk with her in French, stopped calling her \textit{my dear} and using the formal \textit{you}, and invited her to bring all her children to stay the night, to which the princess agreed and, sitting a little longer, went home.

So many guests arrived that day with congratulations that all morning, in the courtyard near the porch, there were several carriages parked at all times.

\textit{``Bonjour, ch\`ere cousine,''} said one of the guests, entering the room and kissing grandmother's hand. %princeivan

He was a man of about seventy, tall, in a military uniform with large epaulets, at the collar of which a large white cross could be seen, the expression on his face calm and open. I found the freedom and simplicity of his movements striking. Despite the fact that he had only a half-circle of thinning hair around the crown of his head and the set of his upper lip clearly demonstrated a lack of teeth, his face was remarkably good-looking.

Prince Ivan Ivanych, at the end of the previous century,\todo{misplaced} thanks to his noble character, attractive physical appearance, remarkable bravery, noble and strong blood, and especially to good fortune, made a brilliant career at a very young age. He continued to serve and very quickly his ambition was satisfied, such that he could wish for nothing further in that regard. From early youth, he held himself as if ready to occupy the very same brilliant position in society in which fate subsequently placed him; consequently, although in his brilliant and vain life, as in all others, he met with misfortunes, disappointments, and afflictions, he never once betrayed his lofty way of thinking, or his calm character, or the fundamental rules of religion and morality, and he gained the esteem of others not so much on the basis of his brilliant position as on the basis of his consistency and steadfastness. He was no great mind, but, thanks to his position, which allowed him to look on all the vain troubles of life from above, his way of thinking was elevated. He was kind and sensitive, but cold and somewhat haughty in his behavior. This occurred because, having been placed in a position where he could be useful to many, through coldness he sought to protect himself from incessant requests and favors by people who only wanted to use his influence. This coldness was softened, however, by the condescending courteousness of a man of \emph{very great society}. He was educated and well read; but his education stopped at what he had acquired in his youth, that is, at the end of the previous century. He had read everything that had been written during the 18th century in France, a time known for philosophy and eloquence, knew thoroughly all the best works of French literature, meaning he could (and loved) to cite frequently passages from Racine, Corneille, Boileau, Moliere, Montaigne, Fenelon; he possessed brilliant knowledge of mythology had profitably studied, in French translations, the ancient masterpieces of epic poetry, possessed satisfactory knowledge of history, obtained from Segur; but he possessed no understanding of mathematics further than arithmetic, nor of physics, nor of contemporary literature: in conversation he could hold silent decently or say a few general phrases about Goethe, Schiller, or Byron, but he had never read them. Despite this French-classical education, of which there now remain very few examples, his conversation was simple, and this simplicity equally hid his ignorance of a few things and set a pleasant tone and tolerance of others. He was a great enemy of every kind of originality, saying that originality was a trick for people of a bad sort. Society was essential for him, wherever he lived; in Moscow or abroad, he always lived with identical openness and on certain days entertained the entire city. He was on such terms with everyone in the city that an invitation card from him could serve as a passport into every drawing room, many young and beautiful ladies eagerly presented their rosy cheeks to him to be kissed with a kind of fatherly feeling, and others, apparently important and upstanding people, felt indescribable pleasure when invited by the prince.

Few people like grandmother remained for the prince, who were of the same circle as him, with a similar upbringing, views on things, and of the same age; therefore he especially valued his ancient, friendly ties with her and always showed her the greatest esteem.

I could not get enough of looking at the prince: the esteem that everyone showed him, the large epaulets, the particular joy that grandmother expressed upon seeing him, and the fact that he alone, apparently, did not fear her, conversed with her perfectly freely, and even had the courage to call her \textit{ma cousine}, inspired in me esteem for him that equalled what I felt for grandmother. When shown my verses, he called me over and said:

``Who can tell, \textit{ma cousine}, perhaps he will be another Derzhavin.''\footnote{Gavrila Romanovich Derzhavin (1743--1816) was a highly-esteemed classical Russian poet of the 18th and early 19th centuries.} %princeivan

At this, he pinched my cheek so painfully that, if I did not cry out, it was only because I had managed to guess that it was kindly meant.

The guests all began to leave, papa and Volodya stepped out: in the drawing room, the prince, grandmother, and I remained.

``Why did your lovely Natalya Nikolayevna not come?'' Prince Ivan Ivanych suddenly asked, after a minute's silence. %princeivan

``Ah, \textit{mon cher},'' grandmother answered, lowering her voice and laying her hand on the sleeve of his uniform, ``she most likely would have come had she been free to do as she wished. She wrote to me that Pierre suggested to her that they come, but that she decided against it herself, because they evidently had no income at all this year; and she writes, `Besides, I have no reason to bring the whole household to Moscow this year. Lyubochka is still too small; and as far as the boys, who are coming to live with you, are concerned, I am much more at ease than if they were with me.' Isn't that splendid!'' grandmother continued in a tone that clearly demonstrated that she did not find it splendid at all. ``It was long since time to send the boys here, so that they can learn something and get used to society; what kind of education could they get in the country, after all? The oldest is thirteen years old, and the other eleven. You may have noticed, \textit{mon cousin}, that they are absolutely wild\ldots{}don't even know how to enter a room.'' %grandmother

``Still, I cannot understand,'' the prince answered, ``why do we hear these constant complaints about distressed circumstances? \emph{He} has a very good situation, and Natasha's Khabarovka, where you and I, in days of yore, played in the theatre, I know it like the back of my hand---a marvelous estate! It should always bring a splendid income.'' %princeivan

``I will tell you, as a true friend,'' grandmother interrupted him, a sad expression on her face. ``It seems to me that these are all just excuses for \emph{him} to live here alone, ramble about the clubs and various dinners, and do God knows what else; and she suspects nothing. You know that angelic kindness---she believes \emph{him} in everything. He assured her that the children needed to be brought to Moscow, but that she should stay in the country with that stupid governess---she believed him; if you told her that children needed to be whipped like Princess Varvara Ilyinichna whips hers and I think she would agree to it on the spot,'' said grandmother, turning in her chair, with a look of utter contempt on her face. ``Yes, my friend,'' grandmother continued after a minute's silence, taking one of her two handkerchiefs in her hand to wipe away a tear that appeared, ``I often think that \emph{he} can neither value nor understand her, and, despite all her kindness and love toward him and the pains she takes to hide her woe---I know all that very well---she cannot be happy with him; and mark my words, if he does not\ldots{}'' %grandmother

Grandmother covered her face with the handkerchief.

\textit{``Eh! ma bonne amie,''} said the prince reproachfully, ``I see you have not become any more sensible---eternally grieving and crying over imagined woe. You should be ashamed! I've known \emph{him} a long time and know him to be an attentive, kind, and splendid husband, and most importantly---the noblest of men, \textit{un parfait honn\^ete homme}.''\footnote{A perfectly honest man. \textit{Fr.}} %princeivan

Having heard this conversation by accident, which I was not supposed to hear, I found my way out of the room on tip-toe, seriously agitated.

\chapter{The Ivins} %c19

``Volodya! Volodya! The Ivins!'' I shouted, looking out the window and seeing three boys in blue bekesha coats with beaver-skin lapels who, following after their dandyish young governor, crossed from the sidewalk on the opposite side of the street from our house.\footnote{A bekesha was a long men's coat with a gathered waist.}

The Ivins were related to us and were almost the same age as us; soon after our arrival in Moscow, we met them and became friends.

The second Ivin---Seryozha---was a dark-looking, curly-haired boy with a hard, turned-up little nose, pretty, very fresh lips, which only rarely covered his upper row of white teeth, protruding slightly, dark-blue, splendid eyes, and an uncommonly pert expression on his face. He never smiled, but either looked perfectly serious or laughed with all his heart, a ringing, distinct, and extremely captivating laugh.\todo{?} I found his original good looks striking from first glance. I felt an insurmountable attraction to him. Just seeing him was enough to make me happy; and for a time all my heart's strength was focused on that one desire: when I had to pass three or four days without seeing him, I began to miss him, and I became sad to the point of tears. All my dreams, sleeping and waking, were of him: lying down to sleep, I hoped to dream of him: closing my eyes, I saw him before me and cherished the apparition as the purest delight. I would not have trusted anyone in the world with those feelings, so highly did I prize them. Perhaps because he was tired of feeling my restless eyes, incessantly turned toward him, or simply did not feel any liking for me, he was noticeably more interested in playing with and talking to Volodya than me; but I was nevertheless satisfied, wanted nothing more, demanded nothing, and was always ready to sacrifice anything for him. Besides the passionate attraction that he inspired in me, his presence excited in me, to no less a degree, another feeling---fear of bothering him, offending him somehow, being unlikeable: perhaps because his face looked haughty, or because, despising my own physical appearance, I overvalued the virtues of beauty, or, what is most likely of all, because I felt as much fear toward him as I did love, a certain sign of love. The first time that Seryozha talked to me, I so lost my composure from this unexpected happiness that I turned white, then red, and could not answer him. He had the bad habit, when he was lost in thought, of fixing his eyes on one point and blinked incessantly, twitching his nose and brows at the same time. Everyone found that this habit spoiled his looks, but I found it so endearing that I unconsciously picked up the habit, and, a few days after I met him, grandmother asked whether my eyes were hurting me, since I was flapping them open and shut like an owl. No word was said between us about love; but he felt his power over me and unknowingly, but tyrannically, used it in our childish relations; I, no matter how much I wished to tell him everything that was on my heart, was too afraid of him to dare to be open; I tried to seem indifferent and submitted to him uncomplainingly. Occasionally, his influence seemed heavy to me, intolerable; but it was not in my power to get out from under him.

I find it sad to remember this fresh, splendid feeling of selfless and limitless love, which died unexpressed and unrequited.

It is strange---I wonder why, when I was a child, I tried to seem like a grown-up, while now that I am one no longer, I often wish I could be like one. So many times that desire---not to seem like a little child in my relations with Seryozha---stopped a feeling that was yearning to find expression and forced me to hide my true feelings. Not only did I not dare to kiss him, which I sometimes wanted to do very much, take him by the hand, say how happy I was to see him, but I did not even dare to call him Seryozha---it was always Sergei: that was how things were between us. Every expression of sensitivity demonstrated childishness and proved me to be a \emph{little boy}. Not having gone through those bitter trials that drive adults to be careful and cold in their relations with others, we denied ourselves the pure pleasure of tender, childish affection only out of a strange desire to pretend to be \emph{grown up}.

I met the Ivins while they were still in the servants' room, greeted them, and then flew back to grandmother: I told her that the Ivins had arrived as if the news should bring her complete happiness in life. Then, not taking my eyes off Seryozha, I went after him into the drawing room and followed his every movement. In the time it took grandmother to say that he had grown a great deal and direct her penetrating eyes at him, I felt the feeling of fear and hope that an artist must experience when awaiting a verdict on his work from an esteemed judge.

The Ivins' young governor, Herr Frost, with grandmother's permission, went with us to the front garden, sat on the green bench there, arranged his legs, the perfect picture of a gentleman, placing his cane, topped with a bronze knob, between them, and, with the look of a person who was very satisfied with his actions, lit a cigar.

Herr Frost was a German, but a German cut from an entirely different cloth than our kind Karl Ivanych: first, he spoke very correct Russian, and French with a very bad accent, and enjoyed the reputation, especially among women, of a very learned person; second, he had a reddish moustache and wore a ruby stick-pin in his black satin scarf, the ends of which were shoved under his suspenders, and light blue pants made of a shot fabric, with trouser stripes; third, he was young, had an attractive, self-satisfied appearance, and uncommonly visible, muscular legs. It was obvious that this last virtue he prized especially: he thought their effects to be irresistible in relation to individuals of the female sex, and it must have been with this goal in mind that he tried to display his legs in the most visible place---standing or sitting, he always brought his calves into motion.\todo{needs work} He was the very type of a young Russian German who wanted to be a good fellow\todo{ugh} and a ladies' man.

We enjoyed ourselves in the front garden. A game of robbers was going very well; but one thing threatened to ruin everything. Seryozha was a robber: chasing after some passing riders, he stumbled and hit his knee against a tree at top speed, so sharply that I thought that he would break into pieces. Despite the fact that I was a gendarme and my duty was to catch him, I approach him and began to ask with concern whether he was hurt. Seryozha got angry at me: he made a fist, stomped his feet, and in a voice that clearly demonstrated that he was badly hurt, shouted at me:

``Well, so what? You've ruined the game now! Well, why aren't you trying to catch me? Why aren't you trying to catch me?'' he repeated several times, looking sidelong at Volodya and the older Ivin, who were pretending to be the passing riders, hopping and running along the path, and he suddenly screamed and ran off to catch them with a loud laugh.

I can hardly relate how amazed, captivated I was by this heroic act: despite the frightful pain, he not only did not cry, but gave no sign that he was hurt, and did not forget the game for a minute.

Soon after this, when Ilyinka Grap had also joined our company, and we were going upstairs to dinner, Seryozha had another chance to captivate and amaze me with his astonishing fortitude and strength of character.

Ilyinka Grap was the son of a poor foreigner who at one point lived in my grandfather's house, was somehow obligated to him, and now considered it his solemn duty to send his son, as frequently as possible, to stay with us. If he supposed that an acquaintance with us would afford his son some kind of honor or pleasure, then he was absolutely mistaken in that regard, because we not only were not friendly with Ilyinka, but we paid attention to him only when we wanted to laugh at him. Ilyinka Grap was a boy about thirteen years old, skinny, tall, pale, with an ugly mug like a bird's, and good-natured, submission expression. He was very poorly dressed, but at the same time pomaded his hair so abundantly that we were certain that on a sunny day Grap's pomade would melt on his head and pour down the back of his jacket. When I remember him now, I find that he was a very obliging, quiet, and kind boy; at the time, he seemed to me such a detestable creature that it was not worth feeling sorry for him or even thinking about him.

When the game of robbers came to an end, we went upstairs, began to \emph{busy ourselves} and show off to one another with various school-boy tricks. Ilyinka looked at us with a timid, surprised smile, and when we proposed that he give it a try, he refused, saying that he did not have the strength. Seryozha was surprisingly nice; he took off his jacket---his face and eyes burned---he roared with laughter incessantly and came up with new pranks to play: he jumped over three chairs placed in a row, rolled a wheel through the entire room, stood on his head on top of copies of Tatishchev's lexicon, which we had placed in the middle of the room as a kind of pedestal, and at the same time played tricks with his legs that killed us, making it impossible for us to hold back from laughing. After the last trick, he thought for a minute, blinked his eyes, and suddenly, with an absolutely serious face, went up to Ilyinka: ``try to do it; it's really not hard.'' Grap, noting that everyone's attention was on him, turned red and in a barely audible voice assured him that he could not possibly do it. %seryozha

``What's this all about, why doesn't he want to show us anything? What a girl he is\ldots{} He definitely needs to stand on his head!'' %seryozha

And Seryozha took him by the hand.

``Definitely, definitely, stand him on his head!'' we all shouted, surrounding Ilyinka, who at that moment was frightened and turning white, grabbed him by the hand and pulled him toward the lexicons.

``Leave me alone, I'll do it myself! You'll rip my jacket!'' shouted the unhappy victim. But his shouts of despair further emboldened us; we were dying of laughter; the green jacket was straining at every seam.

Volodya and the eldest Ivin pushed his head down and placed him on the lexicons; Seryozha and I grabbed the poor boy by his thin little legs, which he waved in various directions, rolled up his pant legs to the knees, and, with loud laughter, flung them upwards; the youngest Ivin kept his whole body in balance.

It so happened that, after some loud laughter, we all suddenly fell silent, and the room became so quiet that only the unhappy Grap's heavy breathing could be heard. At that moment, I was not entirely convinced that everything was happy and funny.

``Now that is a good fellow,'' said Seryozha, giving him a slap of the hand. %seryozha

Ilyinka was silent and, trying to pull away, threw his legs in various directions. In one of these desperate movements, he hit Seryozha in the eye with his heel so hard that Seryozha let go of his legs right away, grabbed his eye, from which tears were flowing involuntarily, and pushed Ilyinka with all his strength. Ilyinka, no longer held up by us, fell in a kind of lifeless way to the ground and through his tears could only get out:

``What are you tormenting me for?'' %ilyinka

The pitiable figure of poor Ilyinka, with his tear-stained face, disheveled hair, and rolled-up pants with unpolished boot-tops visible underneath, was striking to us; we were all silent and tried to force smiles.

Seryozha came to his senses first.

``What an old woman, a whiner,'' he said, touching him gently with his foot. ``You can't take a joke\ldots{} Well, enough, get up.'' %seryozha

``I already told you that you're nasty,'' Ilyinka said angrily, and, turning away, he began to sob loudly. %ilyinka

``Oh! Now he's gonna\todo{?} stomp his feet, cry, and moan!'' shouted Seryozha, grabbing a lexicon in his hands and waving it over the head of the unhappy boy, who did not even think to protect himself and just covered his head with his hands.

``Here you go! Here you go! Let's drop him if he can't take a joke\ldots{} Let's go downstairs,'' said Seryozha, laughing unnaturally. %seryozha

I looked at the poor young man with concern; he was lying on the floor and hiding his face in the lexicons, crying so much that it seemed he might die of the convulsions that shook his whole body.

``Hey, Seryozha!'' I said to him, ``why did you do that?'' %nikolai

``Here's another one! I didn't start crying today when I hit my leg and almost broke the bone, did I?'' %seryozha

``Yes, that's true,'' I thought. ``Ilyinka is nothing more than a crybaby, and Seryozha here---he's a good fellow\ldots{}what a good fellow!'' %nikolai

I could not even imagine that the poor young man was probably crying not so much from physical paint as from the thought that five boys, who he might even have liked, decided without any reason to hate him and persecute him.

I absolutely could not explain to myself the cruelty of my own actions. Why did I not go up to him, not protect him, and not comfort him? Where did my feelings of sympathy go, which would sometimes cause me to cry my eyes out at the sight of a baby jackdaw thrown out of its nest, or a puppy that someone was carrying to throw over the fence, or a chicken that the cook's assistant was carrying in for soup?

Could those splendid feelings really be drowned out by my love for Seryozha and my desire to seem to be a good fellow like he was? How despicable were that love and that desire to be a good fellow! They caused the only black spots on the pages of my memories of childhood.

\chapter{The Guests Arrive} %c20
